---
title: "Survival Simulation — Stage 1: Main Loop & Aggregation"
subtitle: "Comparing Cox vs Random Survival Forest"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# Background

This document runs **Stage 1**: the 100-replication Monte Carlo loop and the
final comparison. It loads `setup.rds` from Stage 0 (which fixes $\Sigma$,
$\theta$, the RSF hyperparameters, and the time grid) and reuses the same DGP.

Each replication does the following:

1. Generate a fresh **training** sample of $n=5000$: $(X_i, T_i, \Delta_i)$.
2. Fit **Cox** and **RSF**.
3. Generate a fresh **test** sample of $m=20$ covariate vectors.
4. Evaluate each model's $\hat S(t\mid x)$ and the **true** $S(t\mid x)$ on the
   $m \times M$ grid.
5. Compute $\mathrm{MAD}_{\text{Cox}}$ and $\mathrm{MAD}_{\text{RSF}}$ and store the pair.

The replications are **run in parallel** across the machine's cores, then
aggregated to compare the two estimators.

```{r libs}
suppressPackageStartupMessages({
  library(MASS)             # mvrnorm
  library(Matrix)           # nearPD (only if rebuilding Sigma)
  library(survival)         # coxph, basehaz
  library(randomForestSRC)  # rfsrc, predict
  library(parallel)         # makeCluster, parLapply
})
```

## Load the Stage 0 setup

```{r load-setup}
setup <- readRDS("setup.rds")

MASTER_SEED <- setup$meta$master_seed
BETA        <- setup$beta
A0          <- 0.005

N_TRAIN <- 5000L
M_TEST  <- 20L
N_REP   <- 100L     # set to e.g. 5 first for a fast smoke test, then 100

str(setup[c("theta", "rsf_hp", "meta")])
cat(sprintf("\ntime grid: M=%d, range (%.3f, %.3f]\n",
            length(setup$time_grid), min(setup$time_grid), max(setup$time_grid)))
```

# DGP functions (reused from Stage 0)

Identical to Stage 0 — restated here so this document runs standalone.

```{r dgp}
sample_X <- function(n, Sigma) {
  MASS::mvrnorm(n = n, mu = rep(0, ncol(Sigma)), Sigma = Sigma)
}

sample_event_times <- function(X, beta = BETA) {
  lp <- as.vector(X %*% beta)
  U  <- runif(nrow(X))
  sqrt(-log(U) / (A0 * exp(lp)))
}

make_survival_sample <- function(n, Sigma, theta, beta = BETA) {
  X      <- sample_X(n, Sigma)
  Ttilde <- sample_event_times(X, beta)
  C      <- rexp(n, rate = 1 / theta)
  Tobs   <- pmin(Ttilde, C)
  Delta  <- as.integer(Ttilde <= C)
  colnames(X) <- paste0("x", seq_len(ncol(X)))
  data.frame(X, time = Tobs, status = Delta)
}
```

# Estimating $\hat S(t\mid x)$ on the grid

Both estimators are **step functions** defined on their own event times, so we
evaluate them at our fixed grid via `findInterval` (the cumulative value at the
largest event time $\le t$). This keeps the time axis identical across models
and against the truth.

## Cox (Breslow baseline)

The estimated cumulative hazard factorizes as
$\hat\Lambda(t\mid x)=\hat\Lambda_0(t)\,e^{\hat\beta^\top x}$, where
$\hat\Lambda_0$ is the Breslow baseline. We pull $\hat\Lambda_0$ from
`basehaz(fit, centered = FALSE)` (uncentered, so the linear predictor is exactly
$\hat\beta^\top x$) and form
$\hat S(t\mid x)=\exp\!\big(-\hat\Lambda_0(t)\,e^{\hat\beta^\top x}\big)$.
Returns an $m \times M$ matrix.

```{r est-cox}
cox_survival_matrix <- function(fit, Xtest, tgrid) {
  bh  <- basehaz(fit, centered = FALSE)              # columns: hazard (cum), time
  idx <- findInterval(tgrid, bh$time)                # H0 step evaluation
  H0  <- ifelse(idx == 0, 0, bh$hazard[idx])         # length M
  b   <- coef(fit)
  lp  <- as.vector(Xtest[, names(b), drop = FALSE] %*% b)  # length m
  exp(-outer(exp(lp), H0))                           # m x M : S(t_i | x_j)
}
```

## RSF

`predict.rfsrc` returns a survival matrix over `time.interest` (the unique event
times). We evaluate at our grid the same way; before the first event time
$\hat S = 1$. Returns an $m \times M$ matrix.

```{r est-rsf}
rsf_survival_matrix <- function(fit, Xtest_df, tgrid) {
  pr   <- predict(fit, newdata = Xtest_df)
  ti   <- pr$time.interest                           # increasing, length K
  surv <- pr$survival                                # m x K
  idx  <- findInterval(tgrid, ti)                    # length M
  out  <- matrix(1, nrow = nrow(surv), ncol = length(tgrid))  # S=1 before t1
  nz   <- idx > 0
  if (any(nz)) out[, nz] <- surv[, idx[nz], drop = FALSE]
  out                                                # m x M
}
```

## True survival and the MAD metric

```{r est-true}
true_survival_matrix <- function(Xtest, tgrid, beta = BETA) {
  lp <- as.vector(Xtest %*% beta)
  exp(-A0 * outer(exp(lp), tgrid^2))                 # m x M : true S(t_i | x_j)
}

# MAD = (1 / (m*M)) * sum |Shat - Strue|  = mean over all m*M cells
mad_metric <- function(Shat, Strue) mean(abs(Shat - Strue))
```

# One replication

Each replication is fully reproducible via a per-rep seed
(`MASTER_SEED + 1000 + rep_id`) that controls both the data generation
(R's RNG) and the RSF fit (passed to `rfsrc` as a negative seed), so any single
rep can be re-run in isolation.

```{r run-one}
run_one <- function(rep_id, setup) {
  set.seed(MASTER_SEED + 1000L + rep_id)

  # 1. training sample
  train <- make_survival_sample(N_TRAIN, setup$Sigma, setup$theta, setup$beta)

  # 2. fit models
  cox <- coxph(Surv(time, status) ~ ., data = train)
  rsf <- rfsrc(Surv(time, status) ~ ., data = train,
               ntree    = setup$rsf_hp$ntree,
               mtry     = setup$rsf_hp$mtry,
               nodesize = setup$rsf_hp$nodesize,
               seed     = -(MASTER_SEED + 1000L + rep_id))  # rfsrc needs a negative seed

  # 3. fresh test covariates
  Xtest <- sample_X(M_TEST, setup$Sigma)
  colnames(Xtest) <- paste0("x", seq_len(ncol(Xtest)))
  Xtest_df <- as.data.frame(Xtest)

  # 4. truth + estimates on the grid
  Strue <- true_survival_matrix(Xtest,    setup$time_grid, setup$beta)
  Scox  <- cox_survival_matrix(cox,       Xtest,           setup$time_grid)
  Srsf  <- rsf_survival_matrix(rsf,       Xtest_df,        setup$time_grid)

  # 5. MAD per model
  c(mad_cox = mad_metric(Scox, Strue),
    mad_rsf = mad_metric(Srsf, Strue))
}
```

# The 100-replication loop (parallel)

Replications are independent, so we distribute them across the machine's cores.
Each worker runs `rfsrc` single-threaded (`rf.cores = 1`) to avoid
oversubscription — since process-level parallelism here is what actually gives
the speedup. Results are identical to a serial run because every replication
seeds itself from `rep_id`, so execution order is irrelevant.

The `cache.extra` key ties this chunk's cache to `setup.rds`, so changing the
setup (e.g. new RSF hyperparameters) correctly forces a re-run.

```{r loop, cache.extra = file.mtime("setup.rds")}
n_cores <- max(1L, detectCores() - 1L)   # leave one core for the OS
cl <- makeCluster(n_cores)

# prepare workers: load packages, disable nested rfsrc threading
clusterEvalQ(cl, {
  suppressPackageStartupMessages({
    library(MASS); library(survival); library(randomForestSRC)
  })
  options(rf.cores = 1, mc.cores = 1)
  NULL
})

# ship functions + globals to the workers
clusterExport(cl,
  varlist = c("run_one", "make_survival_sample", "sample_X",
              "sample_event_times", "cox_survival_matrix",
              "rsf_survival_matrix", "true_survival_matrix", "mad_metric",
              "MASTER_SEED", "BETA", "A0", "N_TRAIN", "M_TEST", "setup"),
  envir = environment())

# run in core-sized batches so we can print progress + ETA
batches <- split(seq_len(N_REP), ceiling(seq_len(N_REP) / n_cores))
results <- vector("list", N_REP)
t0 <- Sys.time(); done <- 0L

cat(sprintf("running %d reps on %d cores...\n", N_REP, n_cores))
for (b in batches) {
  results[b] <- parLapply(cl, b, run_one, setup = setup)
  done <- done + length(b)
  el <- as.numeric(difftime(Sys.time(), t0, units = "mins"))
  cat(sprintf("  %3d/%d reps  (%.1f min elapsed, ~%.1f min left)\n",
              done, N_REP, el, el / done * (N_REP - done)))
}

stopCluster(cl)

results <- as.data.frame(do.call(rbind, results))
saveRDS(results, "results.rds")
```

# Aggregation and comparison

## Summary

```{r agg-summary}
summary_tbl <- data.frame(
  model = c("Cox", "RSF"),
  mean  = c(mean(results$mad_cox), mean(results$mad_rsf)),
  sd    = c(sd(results$mad_cox),   sd(results$mad_rsf)),
  median = c(median(results$mad_cox), median(results$mad_rsf))
)
summary_tbl
```

## Plot

```{r agg-plot, fig.width=5, fig.height=4}
boxplot(results, names = c("Cox", "RSF"),
        col = c("#9ecae1", "#fdae6b"),
        ylab = "MAD", main = "MAD distribution (100 reps)")
```

# What we have

`results.rds` holds the 100 paired MAD values for each model. The summary table
and boxplot give the head-to-head comparison between Cox and RSF: the model with
the lower mean MAD is closer to the true survival curves on this DGP.
