---
title: "Survival Simulation — Stage 0: One-Time Setup (DGP, Calibration, Tuning)"
subtitle: "Comparing Cox vs Random Survival Forest"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,        # RSF tuning is slow — cache avoids re-running on re-knit
  fig.align = "center"
)
options(digits = 4)
```

# Background

This document carries out **Stage 0** of the project: all the one-time
preparations that stay fixed across the 200 Monte Carlo replications. The
end goal is to compare two estimators of the survival function
$S(t\mid X)$ — **Cox PH** vs **Random Survival Forest (RSF)** — on data we
generate from a known model, so that the *true* $S$ is known analytically
and serves as ground truth.

The four outputs of this stage:

1. **$\Sigma$** — a correlation matrix (PD, weak dependence) for sampling covariates.
2. **$\theta$** — the censoring parameter, calibrated to a censoring rate of $\approx 20\%$.
3. **RSF hyperparameters** — tuned via out-of-bag (OOB) error.
4. **Time grid** $t_1,\dots,t_M$ — for evaluating the survival curves.

Everything is saved to `setup.rds` and loaded by the main loop (Stage 1).

```{r libs}
suppressPackageStartupMessages({
  library(MASS)             # mvrnorm
  library(Matrix)           # nearPD
  library(survival)         # coxph
  library(randomForestSRC)  # rfsrc, tune
})
```

## Seeds and global parameters

All sampling is controlled by a single seed, making the run fully
reproducible. Change `MASTER_SEED` to draw a different pilot.

```{r params}
MASTER_SEED   <- 20240601L

# --- DGP constants ---
P    <- 6L
BETA <- c(0.1, 0.1, 0.1, -0.1, -0.1, -0.1)
A0   <- 0.005            # Lambda_0(t) = A0 * t^2  (from lambda_0(t) = 0.01 t)

# --- Simulation hyperparameters ---
N_PILOT       <- 50000L  # large pilot sample for stable calibration/tuning
N_TUNE        <- 5000L   # sample size for RSF tuning (= actual run size)
TARGET_CENSOR <- 0.20    # target censoring rate
M_TIMES       <- 100L    # number of points in the time grid
TIME_QUANTILE <- 0.90    # upper cutoff for the grid
SIGMA_RHO     <- 0.2
SIGMA_TYPE    <- "ar1"   # "ar1" or "cs"
```

# The DGP module

The functions below are the core of the simulation and will be reused in
the main loop. They are pure (no `set.seed` inside) so that randomness is
controlled externally and explicitly.

## Base definitions and building $\Sigma$

The cumulative baseline is derived directly from the definition
$\lambda_0(t)=0.01t$:
$$\Lambda_0(t)=\int_0^t 0.01\,u\,du = 0.005\,t^2.$$
$\Sigma$ is built as AR(1) or compound-symmetry with a small $\rho$, and
projected to PD if needed.

```{r dgp-base}
Lambda0 <- function(t) A0 * t^2

build_sigma <- function(p = P, rho = SIGMA_RHO, type = SIGMA_TYPE) {
  type <- match.arg(type, c("ar1", "cs"))
  S <- if (type == "cs") {
    M <- matrix(rho, p, p); diag(M) <- 1; M
  } else {
    rho ^ abs(outer(seq_len(p), seq_len(p), "-"))
  }
  if (min(eigen(S, symmetric = TRUE, only.values = TRUE)$values) <= 1e-8) {
    S <- as.matrix(Matrix::nearPD(S, corr = TRUE)$mat)
  }
  S
}
```

## Sampling covariates and event times

The theoretical core is **Inverse Transform Sampling**. From the model,
$$S(t\mid X)=\exp\!\big(-A_0\,t^2\,e^{\beta^\top X}\big).$$
Using the property $S(\tilde T\mid X)\sim U(0,1)$, we set
$S(\tilde T_i\mid X_i)=U_i$ and solve for $\tilde T_i$:
$$\tilde T_i=\sqrt{\dfrac{-\ln U_i}{A_0\,e^{\beta^\top X_i}}}.$$
Since $-\ln U_i\sim \mathrm{Exp}(1)$, this is a closed form — no numerical
solve required.

```{r dgp-sample}
sample_X <- function(n, Sigma) {
  MASS::mvrnorm(n = n, mu = rep(0, ncol(Sigma)), Sigma = Sigma)
}

# True event times T~_i  (inverse-transform formula)
sample_event_times <- function(X, beta = BETA) {
  lp <- as.vector(X %*% beta)
  U  <- runif(nrow(X))
  sqrt(-log(U) / (A0 * exp(lp)))
}

# True survival function S(t | x): x a single covariate vector, t a vector of times
true_survival <- function(t, x, beta = BETA) {
  exp(-A0 * t^2 * exp(sum(x * beta)))
}
```

## Censoring and assembling a sample

The censoring time $C\sim\mathrm{Exp}$ with **mean $=\theta$** (i.e.
rate $=1/\theta$), independent of $\tilde T$. The observed datum:
$$T_i=\min(\tilde T_i,C_i),\qquad \Delta_i=\mathbb 1(\tilde T_i\le C_i).$$

For calibration we exploit that $C\perp\tilde T$ lets us integrate out $C$
analytically, given the pilot's $\tilde T$ values:
$$g(\theta)=\mathbb P(C<\tilde T)=\frac1n\sum_i\Big(1-e^{-\tilde T_i/\theta}\Big).$$
This makes calibration deterministic — no sampling noise.

```{r dgp-censor}
# Analytic censoring rate as a function of theta (mean), given the pilot's T~
censoring_rate <- function(theta, Ttilde) mean(1 - exp(-Ttilde / theta))

# Full survival sample (X, time, status)
make_survival_sample <- function(n, Sigma, theta, beta = BETA) {
  X      <- sample_X(n, Sigma)
  Ttilde <- sample_event_times(X, beta)
  C      <- rexp(n, rate = 1 / theta)      # Exp with mean = theta
  Tobs   <- pmin(Ttilde, C)
  Delta  <- as.integer(Ttilde <= C)        # 1 = event, 0 = censored
  colnames(X) <- paste0("x", seq_len(ncol(X)))
  data.frame(X, time = Tobs, status = Delta)
}
```

# Stage 0.1 — Correlation matrix $\Sigma$

```{r sigma}
Sigma <- build_sigma()
stopifnot(min(eigen(Sigma, symmetric = TRUE, only.values = TRUE)$values) > 0)

cat(sprintf("Sigma %dx%d (%s, rho=%.2f) — PD OK\n",
            nrow(Sigma), ncol(Sigma), SIGMA_TYPE, SIGMA_RHO))
round(Sigma, 3)
```

# Stage 0.2 — Calibrating $\theta$ to 20% censoring

$g(\theta)$ is **monotonically decreasing** in $\theta$ (larger mean →
later censoring → less censoring), so `uniroot` on $g(\theta)-0.20$
converges to a unique root. After calibration we verify empirically on a
**separate** sample where $C$ is actually drawn.

```{r theta}
set.seed(MASTER_SEED)

X_pilot      <- sample_X(N_PILOT, Sigma)
Ttilde_pilot <- sample_event_times(X_pilot)

calib <- uniroot(
  f        = function(th) censoring_rate(th, Ttilde_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 10 * max(Ttilde_pilot)),
  tol      = 1e-6
)
theta_hat <- calib$root

# Empirical check (separate sample, C actually drawn)
chk_df     <- make_survival_sample(N_PILOT, Sigma, theta_hat)
emp_censor <- mean(chk_df$status == 0)

cat(sprintf("theta = %.4f\n  censoring  analytic = %.3f | empirical = %.3f\n",
            theta_hat,
            censoring_rate(theta_hat, Ttilde_pilot),
            emp_censor))
```

# Stage 0.3 — Tuning RSF (OOB)

`tune()` scans the $(\texttt{nodesize},\texttt{mtry})$ space and minimizes
OOB error — these two have an interior optimum, so tuning them is meaningful.
`ntree` is different: OOB error decreases monotonically in the number of trees
and then flattens, so there is no minimum to find. Since more trees never hurt
accuracy (only compute), we fix `ntree = 1000` — safely past the plateau.

```{r rsf-tune}
set.seed(MASTER_SEED + 1L)
df_tune <- make_survival_sample(N_TUNE, Sigma, theta_hat)

rsf_tune <- tune(
  Surv(time, status) ~ .,
  data      = df_tune,
  ntree.try = 300,             # small forests for fast tuning sweeps
  do.best   = FALSE,           # we build the working forest ourselves
  seed      = MASTER_SEED      # deterministic tuning
)

rsf_hp <- list(
  ntree    = 1000L,
  mtry     = as.integer(rsf_tune$optimal["mtry"]),
  nodesize = as.integer(rsf_tune$optimal["nodesize"])
)

cat(sprintf("RSF tuned (OOB): mtry=%d, nodesize=%d, ntree=%d\n",
            rsf_hp$mtry, rsf_hp$nodesize, rsf_hp$ntree))
```

# Stage 0.4 — The time grid

A uniform grid of $M=100$ points from (0) up to the 90th percentile of the
follow-up times. The cutoff prevents artificial tail error from the step
estimators flattening out after the last event time.

```{r grid}
upper     <- as.numeric(quantile(df_tune$time, TIME_QUANTILE))
time_grid <- seq(0, upper, length.out = M_TIMES + 1L)[-1L]  # M points > 0

cat(sprintf("time grid: M=%d, range (%.3f, %.3f]\n",
            length(time_grid), min(time_grid), max(time_grid)))
```

# Sanity checks

Before saving, verify the DGP behaves as expected.

**(a) Inverse-transform formula.** If sampling is correct, then
$A_0\,\tilde T_i^2\,e^{\beta^\top X_i}=-\ln U_i\sim\mathrm{Exp}(1)$ — i.e.
mean $\approx 1$ and variance $\approx 1$.

```{r check-inv}
lp_chk   <- as.vector(X_pilot %*% BETA)
exp1_chk <- A0 * Ttilde_pilot^2 * exp(lp_chk)
cat(sprintf("Exp(1) check: mean=%.3f (~1), var=%.3f (~1)\n",
            mean(exp1_chk), var(exp1_chk)))
```

**(b) Monotonicity of $S$.** The true survival function decreases in $t$
and lies in $[0,1]$.

```{r check-surv}
s_chk <- true_survival(time_grid, x = X_pilot[1, ])
cat(sprintf("S range [%.3f, %.3f]; monotone decreasing: %s\n",
            min(s_chk), max(s_chk), all(diff(s_chk) <= 0)))
```

**(c) $\Sigma$ — empirical vs theoretical correlation.**

```{r check-sigma}
cat(sprintf("max |empirical corr - Sigma| = %.3f\n",
            max(abs(cor(X_pilot) - Sigma))))
```

# Saving the setup

```{r save}
setup <- list(
  Sigma     = Sigma,
  theta     = theta_hat,
  rsf_hp    = rsf_hp,
  time_grid = time_grid,
  beta      = BETA,
  meta      = list(
    n_pilot       = N_PILOT,
    target_censor = TARGET_CENSOR,
    emp_censor    = emp_censor,
    sigma_type    = SIGMA_TYPE,
    sigma_rho     = SIGMA_RHO,
    master_seed   = MASTER_SEED
  )
)
saveRDS(setup, "setup.rds")
cat("setup.rds saved — ready for the main loop (Stage 1).\n")
```

---

**Next up:** Stage 1 — the 200-replication loop (fitting Cox and RSF,
computing MAD against the truth) and aggregation. It will load `setup.rds`
and reuse the same DGP functions.
