---
title: "Survival Simulation 02 — Stage 0: Setup (DGP, Calibration, Tuning)"
subtitle: "Gompertz-shape hazard DGP (paper Sec 7.1) with left truncation"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# Background

This is the second simulation in this project, reusing the format established
by simulation 01 (`../01_cox_random_forerst_no_lt/`) — same two-stage
structure, same shared engine (`../_common/sim_engine.R`), same reporting
(`../_common/sim_report.R`) — but on a different data-generating process.

The DGP is taken from Section 7.1 of "Flexible Deep Neural Networks for
Partially Linear Survival Data" (`../../papers/Survival-Deep2512.10570v2.pdf`).
That paper compares two deep-learning estimators; we take only their DGP, and
fit the same Cox and RSF arms used in simulation 01. Left truncation is not
part of the paper — it is added here the same way simulation 01 adds it to its
own DGP.

## Why a second DGP

Simulation 01's DGP is an exactly-Cox proportional-hazards model, so the Cox
arms are correctly specified there. Here the hazard is
$$h(t\mid X,Z) = 0.1\exp\{a(X)\,t + \theta_0^\top Z\}, \qquad a(X)=0.1+f(X)^2,$$
with $f(X) = 0.2(X_1+X_2) + 0.5X_1X_2 + X_3^2$ nonlinear in the nuisance
covariates $X=(X_1,X_2,X_3)$, and $Z=(Z_1,Z_2)$ the primary covariates with
$\theta_0=(2,-1)$. Because $a(X)$ multiplies $t$ inside the exponent, the
hazard's *shape* over time depends on $X$, not just its level — this is not
proportional hazards jointly in $(X,Z)$, so a Cox fit here is misspecified for
the nuisance effect by construction. Running the same left-truncation question
on a structurally different DGP is the point of a second simulation.

```{r libs}
source("dgp.R")
suppressPackageStartupMessages({
  library(survival)
  library(randomForestSRC)
})
```

The cumulative hazard has closed form,
$$H(t\mid X,Z) = 0.1\,e^{\theta_0^\top Z}\,\frac{e^{a(X)t}-1}{a(X)},\qquad
S(t\mid X,Z)=\exp(-H(t\mid X,Z)),$$
and inverse-transform sampling follows the same closed form used in
simulation 01: with $E\sim\mathrm{Exp}(1)$,
$$\tilde T = \frac{1}{a(X)}\log\!\Big(1+\frac{a(X)E}{0.1\,e^{\theta_0^\top Z}}\Big).$$

## A discrepancy in the paper's censoring spec

Section 7.1 states an independent exponential censoring time "with mean
$1/30$", together with administrative censoring at the study horizon
$\tau=30$, "resulting censoring rate ... approximately 30%." Neither literal
reading of "mean $1/30$" reproduces 30%: simulated at $n=200{,}000$,
mean $=1/30$ gives $\approx99\%$ censored, and rate $=1/30$ (mean $=30$) gives
$\approx15\%$. We therefore treat **30% as the design target** and calibrate
the censoring parameter to it by root-find, the same approach simulation 01
uses for its own censoring rate — deterministic, no sampling noise, because
censoring is independent of the event time.

Outputs of this stage, all written to `setup.rds`:

1. $\theta$ — right-censoring parameter for the untruncated reference DGP,
   calibrated to 30% censoring (with administrative censoring at $\tau=30$
   folded into the calibration).
2. $c_U$ — entry-window upper bound, calibrated so that $\approx\tfrac13$ of
   the population is truncated away (same target as simulation 01, added here
   rather than in the paper).
3. $\theta_{LT}$ — censoring parameter under truncation, calibrated to 30%
   among the *retained* subjects.
4. Time grid.
5. RSF hyperparameters, tuned by out-of-bag error.

```{r params}
MASTER_SEED   <- 20240815L   # distinct from simulation 01's seed

N_PILOT       <- 50000L
TARGET_CENSOR <- 0.30        # this paper's target, per the note above
TARGET_TRUNC  <- 1/3         # same target as simulation 01
M_TIMES       <- 100L
TIME_QUANTILE <- 0.90

N_POOL        <- 50000L
N_LTRC        <- 5000L       # matches simulation 01's run size
C_L           <- 0
```

# Censoring parameter $\theta$ (untruncated reference)

```{r theta}
set.seed(MASTER_SEED)

XZ_pilot     <- sample_XZ(N_PILOT)
Ttilde_pilot <- sample_event_times(XZ_pilot$X, XZ_pilot$Z)

theta_hat <- uniroot(
  f        = function(th) censoring_rate(th, Ttilde_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 10 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

chk_df     <- make_survival_sample(N_PILOT, theta_hat)
emp_censor <- mean(chk_df$status == 0)

cat(sprintf("theta = %.4f\n  censoring  analytic = %.3f | empirical = %.3f\n",
            theta_hat, censoring_rate(theta_hat, Ttilde_pilot), emp_censor))
cat(sprintf("  P(T~ > tau=%d) = %.4f  (administrative censoring is a minor contributor)\n",
            TAU, mean(Ttilde_pilot > TAU)))
```

# Entry window $c_U$

```{r cU}
cU_hat <- uniroot(
  f        = function(cu) trunc_rate(cu, Ttilde_pilot, C_L) - TARGET_TRUNC,
  interval = c(C_L + 1e-6, 50 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("cU = %.4f  (entry window: Unif(%.2f, %.2f);  cU / tau = %.3f)\n",
            cU_hat, C_L, cU_hat, cU_hat / TAU))
cat(sprintf("  truncated away  analytic = %.3f  (target %.3f)\n",
            trunc_rate(cU_hat, Ttilde_pilot, C_L), TARGET_TRUNC))
```

$c_U \ll \tau$, so entry times fall well inside the study horizon and
administrative censoring plays essentially no role in the truncation
mechanism.

# Censoring under truncation, $\theta_{LT}$

```{r theta-lt}
set.seed(MASTER_SEED + 2L)

R_pilot  <- runif(N_PILOT, C_L, cU_hat)
kept     <- R_pilot <= Ttilde_pilot
Tk_pilot <- Ttilde_pilot[kept]
Rk_pilot <- R_pilot[kept]

theta_lt_hat <- uniroot(
  f        = function(th) censoring_rate_lt(th, Tk_pilot, Rk_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 50 * max(Tk_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("retained %d / %d pool subjects (%.1f%%)\n",
            sum(kept), N_PILOT, 100 * mean(kept)))
cat(sprintf("theta_LT = %.4f\n  censoring (retained)  analytic = %.3f\n",
            theta_lt_hat, censoring_rate_lt(theta_lt_hat, Tk_pilot, Rk_pilot)))
```

# The left-truncated pilot sample

```{r ltrc-pilot}
set.seed(MASTER_SEED + 3L)
df_pilot <- make_ltrc_sample(N_LTRC, N_POOL, theta_lt_hat, cU_hat, C_L)

cat(sprintf("LTRC pilot: n=%d,  censoring=%.3f,  entry <= time on all rows: %s\n",
            nrow(df_pilot), mean(df_pilot$status == 0),
            all(df_pilot$entry <= df_pilot$time)))
```

# Time grid

As in simulation 01, taken from the **left-truncated** pilot — the regime
Stage 1 fits — up to the 90th percentile of observed follow-up.

```{r grid}
upper     <- as.numeric(quantile(df_pilot$time, TIME_QUANTILE))
time_grid <- seq(0, upper, length.out = M_TIMES + 1L)[-1L]

cat(sprintf("time grid: M=%d, range (%.3f, %.3f]\n",
            length(time_grid), min(time_grid), max(time_grid)))
cat(sprintf("cU = %.3f sits at grid position %d of %d\n",
            cU_hat, sum(time_grid < cU_hat) + 1L, length(time_grid)))
```

# RSF hyperparameters

Same two choices as simulation 01: tuning runs on the left-truncated pilot
(the regime Stage 1 fits), and is entry-agnostic (`tune()`'s internal holdout
does not subset extra arguments alongside it, and hyperparameters should not
be selected using information only two of the four Stage 1 arms receive).

```{r rsf-tune}
xn <- c("x1", "x2", "x3", "z1", "z2")

rsf_tune <- tune(
  Surv(time, status) ~ .,
  data      = df_pilot[, c(xn, "time", "status")],
  ntree.try = 300,
  do.best   = FALSE,
  seed      = MASTER_SEED
)

rsf_hp <- list(
  ntree    = 1000L,
  mtry     = as.integer(rsf_tune$optimal["mtry"]),
  nodesize = as.integer(rsf_tune$optimal["nodesize"])
)

cat(sprintf("RSF tuned (OOB): mtry=%d, nodesize=%d, ntree=%d\n",
            rsf_hp$mtry, rsf_hp$nodesize, rsf_hp$ntree))
```

# Sanity checks

**(a) Inverse-transform formula.** If sampling is correct then
$0.1\,e^{\theta_0^\top Z}(e^{a(X)\tilde T}-1)/a(X) \sim \mathrm{Exp}(1)$, so
mean and variance should both be $\approx 1$.

```{r check-inv}
aX_chk <- a_shape(XZ_pilot$X)
lp_chk <- as.vector(XZ_pilot$Z %*% THETA0)
exp1_chk <- BASE_RATE * exp(lp_chk) * (exp(aX_chk * Ttilde_pilot) - 1) / aX_chk
cat(sprintf("Exp(1) check: mean=%.3f (~1), var=%.3f (~1)\n",
            mean(exp1_chk), var(exp1_chk)))
```

**(b) Monotonicity of $S$.**

```{r check-surv}
x1row <- cbind(XZ_pilot$X[1, , drop = FALSE], XZ_pilot$Z[1, , drop = FALSE])
colnames(x1row) <- xn
s_chk <- as.vector(true_survival_matrix(x1row, time_grid))
cat(sprintf("S range [%.3f, %.3f]; monotone decreasing: %s\n",
            min(s_chk), max(s_chk), all(diff(s_chk) <= 0)))
```

**(c) Selection is present.** Retained subjects survived to their own entry
time. Unlike simulation 01, survival here depends on both the nuisance shape
$a(X)$ and the linear term $\theta_0^\top Z$, so we check the $Z$-linear part
for a shift; a full check would require conditioning on $X$ as well, which is
not attempted here.

```{r check-selection}
lp_pool <- as.vector(XZ_pilot$Z %*% THETA0)
cat(sprintf("mean theta0'Z   pool = %+.4f | retained = %+.4f\n",
            mean(lp_pool), mean(lp_pool[kept])))
```

As in simulation 01, this shift is not itself the bias mechanism — the
estimand is the conditional $S(t\mid x,z)$, and Stage 1 evaluates on
covariates drawn from the population distribution. What matters is the risk
set: under $R\perp(\tilde T,X,Z)$ the delayed-entry risk set at each $t$ is a
random subsample of the population risk set at $t$.

**(d) Pool headroom.**

```{r check-pool}
cat(sprintf("need %d, expect ~%.0f retained from a pool of %d\n",
            N_LTRC, (1 - TARGET_TRUNC) * N_POOL, N_POOL))
```

# Saving

```{r save}
setup <- list(
  theta0    = THETA0,
  tau       = TAU,
  theta     = theta_hat,
  rsf_hp    = rsf_hp,
  time_grid = time_grid,
  ltrc      = list(
    cL       = C_L,
    cU       = cU_hat,
    theta_lt = theta_lt_hat,
    n_pool   = N_POOL,
    n_ltrc   = N_LTRC
  ),
  meta      = list(
    n_pilot       = N_PILOT,
    target_censor = TARGET_CENSOR,
    emp_censor    = emp_censor,
    target_trunc  = TARGET_TRUNC,
    emp_retained  = mean(kept),
    emp_censor_lt = mean(df_pilot$status == 0),
    grid_source   = "ltrc",
    tuned_on      = "ltrc",
    master_seed   = MASTER_SEED,
    source        = "Survival-Deep2512.10570v2.pdf, Section 7.1"
  )
)
saveRDS(setup, "setup.rds")
cat("setup.rds saved — ready for Stage 1.\n")
```
