---
title: "Survival Simulation — Stage 0: Setup (DGP, Calibration, Tuning)"
subtitle: "Cox PH data-generating process with left truncation"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,        # RSF tuning is slow — cache avoids re-running on re-knit
  fig.align = "center"
)
options(digits = 4)
```

# Background

This document carries out the one-time preparations that stay fixed across
every Monte Carlo replication. Data is generated from a known model, so the
true $S(t\mid X)$ is available analytically and serves as ground truth.

The model is sourced from `dgp.R` rather than restated here, so Stage 0,
Stage 1, and every parallel worker use one definition.

```{r libs}
source("dgp.R")
suppressPackageStartupMessages({
  library(survival)
  library(randomForestSRC)
})
```

Outputs of this stage, all written to `setup.rds`:

1. **$\Sigma$** — covariate correlation matrix (PD, weak dependence).
2. **$\theta$** — right-censoring parameter for the untruncated reference DGP.
3. **$c_U$** — entry-window upper bound, calibrated so that $\approx\tfrac13$
   of the population is truncated away.
4. **$\theta_{LT}$** — censoring parameter under truncation, calibrated to
   $\approx 20\%$ censoring among the *retained* subjects.
5. **Time grid** $t_1,\dots,t_M$ — where survival curves are evaluated.
6. **RSF hyperparameters** — tuned by out-of-bag error.

## Ordering

Items 5 and 6 are derived from a **left-truncated** pilot sample, which is the
data regime Stage 1 actually fits. That means the truncation parameters must be
calibrated first, so the sequence runs $\Sigma \to \theta \to c_U \to
\theta_{LT} \to$ pilot sample $\to$ grid $\to$ tuning.

```{r params}
MASTER_SEED   <- 20240601L

# --- calibration targets ---
N_PILOT       <- 50000L  # pilot size for analytic calibration
TARGET_CENSOR <- 0.20    # target censoring rate
TARGET_TRUNC  <- 1/3     # target fraction of the pool truncated away
M_TIMES       <- 100L    # points in the time grid
TIME_QUANTILE <- 0.90    # grid cutoff, as a quantile of observed follow-up
SIGMA_RHO     <- 0.2
SIGMA_TYPE    <- "ar1"

# --- left-truncation settings ---
N_POOL        <- 50000L  # population pool drawn before the truncation filter
N_LTRC        <- 5000L   # retained sample size (= the Stage 1 run size)
C_L           <- 0       # entry-window lower bound: R ~ Unif(C_L, cU)
```

# Correlation matrix $\Sigma$

```{r sigma}
Sigma <- build_sigma(P, SIGMA_RHO, SIGMA_TYPE)
stopifnot(min(eigen(Sigma, symmetric = TRUE, only.values = TRUE)$values) > 0)

cat(sprintf("Sigma %dx%d (%s, rho=%.2f) — PD OK\n",
            nrow(Sigma), ncol(Sigma), SIGMA_TYPE, SIGMA_RHO))
round(Sigma, 3)
```

# Censoring parameter $\theta$ (untruncated reference)

Because $C \perp \tilde T$, the censoring rate integrates out in closed form
given the pilot's $\tilde T$ values,
$$g(\theta)=\frac1n\sum_i\Big(1-e^{-\tilde T_i/\theta}\Big),$$
which is monotone decreasing in $\theta$. `uniroot` therefore returns a unique
root with no sampling noise.

This parameter belongs to the untruncated DGP. Stage 1 fits only truncated data,
so it is carried in `setup.rds` for reference rather than consumed downstream.

```{r theta}
set.seed(MASTER_SEED)

X_pilot      <- sample_X(N_PILOT, Sigma)
Ttilde_pilot <- sample_event_times(X_pilot)

theta_hat <- uniroot(
  f        = function(th) censoring_rate(th, Ttilde_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 10 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

chk_df     <- make_survival_sample(N_PILOT, Sigma, theta_hat)
emp_censor <- mean(chk_df$status == 0)

cat(sprintf("theta = %.4f\n  censoring  analytic = %.3f | empirical = %.3f\n",
            theta_hat, censoring_rate(theta_hat, Ttilde_pilot), emp_censor))
```

# Entry window $c_U$

Integrating out $R \sim \mathrm{Unif}(c_L, c_U)$ given the pilot's $\tilde T$:
$$\mathbb P(R > \tilde T_i) =
\begin{cases}
1 & \tilde T_i < c_L\\[2pt]
\dfrac{c_U - \tilde T_i}{c_U - c_L} & c_L \le \tilde T_i \le c_U\\[6pt]
0 & \tilde T_i > c_U
\end{cases}$$
increasing in $c_U$, since a later entry window truncates more people.

```{r cU}
cU_hat <- uniroot(
  f        = function(cu) trunc_rate(cu, Ttilde_pilot, C_L) - TARGET_TRUNC,
  interval = c(C_L + 1e-6, 50 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("cU = %.4f  (entry window: Unif(%.2f, %.2f))\n", cU_hat, C_L, cU_hat))
cat(sprintf("  truncated away  analytic = %.3f  (target %.3f)\n",
            trunc_rate(cU_hat, Ttilde_pilot, C_L), TARGET_TRUNC))
```

# Censoring under truncation, $\theta_{LT}$

Calibrated on the **retained** subjects, since that is the population the
models see. With $C_i = R_i + E_i$ and $E_i\sim\mathrm{Exp}(\text{mean}=\theta_{LT})$,
$$\mathbb P(C_i < \tilde T_i)=\mathbb P(E_i < \tilde T_i - R_i)
= 1 - e^{-(\tilde T_i - R_i)/\theta_{LT}},$$
decreasing in $\theta_{LT}$.

```{r theta-lt}
set.seed(MASTER_SEED + 2L)

R_pilot  <- runif(N_PILOT, C_L, cU_hat)
kept     <- R_pilot <= Ttilde_pilot
Tk_pilot <- Ttilde_pilot[kept]
Rk_pilot <- R_pilot[kept]

theta_lt_hat <- uniroot(
  f        = function(th) censoring_rate_lt(th, Tk_pilot, Rk_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 50 * max(Tk_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("retained %d / %d pool subjects (%.1f%%)\n",
            sum(kept), N_PILOT, 100 * mean(kept)))
cat(sprintf("theta_LT = %.4f\n  censoring (retained)  analytic = %.3f\n",
            theta_lt_hat, censoring_rate_lt(theta_lt_hat, Tk_pilot, Rk_pilot)))
```

# The left-truncated pilot sample

One sample at the Stage 1 run size, used for both the grid and the tuning.
Drawing it once keeps those two derived from the same draw.

```{r ltrc-pilot}
set.seed(MASTER_SEED + 3L)
df_pilot <- make_ltrc_sample(N_LTRC, N_POOL, Sigma, theta_lt_hat, cU_hat, C_L)

cat(sprintf("LTRC pilot: n=%d,  censoring=%.3f,  entry <= time on all rows: %s\n",
            nrow(df_pilot), mean(df_pilot$status == 0),
            all(df_pilot$entry <= df_pilot$time)))
```

# Time grid

A uniform grid of $M$ points from just above 0 up to the 90th percentile of
observed follow-up. The cutoff keeps the grid away from the far tail, where
step estimators flatten out after the last event time and contribute error that
reflects the estimator's tail behaviour rather than the effect under study.

The quantile is taken from the **left-truncated** pilot, matching the regime
Stage 1 fits. Follow-up under truncation runs later than in the untruncated
DGP, so a grid built from untruncated data would stop short of the range the
models actually operate over.

```{r grid}
upper     <- as.numeric(quantile(df_pilot$time, TIME_QUANTILE))
time_grid <- seq(0, upper, length.out = M_TIMES + 1L)[-1L]

cat(sprintf("time grid: M=%d, range (%.3f, %.3f]\n",
            length(time_grid), min(time_grid), max(time_grid)))
cat(sprintf("cU = %.3f sits at grid position %d of %d\n",
            cU_hat, sum(time_grid < cU_hat) + 1L, length(time_grid)))
```

# RSF hyperparameters

`tune()` scans $(\texttt{nodesize}, \texttt{mtry})$ and minimises OOB error;
both have an interior optimum. `ntree` does not — OOB error decreases and then
flattens — so it is fixed at 1000, past the plateau.

Two deliberate choices here:

- Tuning runs on the **left-truncated** pilot, the regime Stage 1 fits.
- Tuning is **entry-agnostic**. `tune()` carves out an internal holdout
  (`data[trn, ]`) without subsetting extra arguments alongside it, so an
  `entry.time` vector passed through `...` would misalign with the rows. It is
  also the neutral choice: all four Stage 1 arms share one set of
  hyperparameters, so those hyperparameters should not be selected using
  information that only two of the arms receive.

```{r rsf-tune}
rsf_tune <- tune(
  Surv(time, status) ~ .,
  data      = df_pilot[, c(grep("^x", names(df_pilot), value = TRUE),
                           "time", "status")],
  ntree.try = 300,
  do.best   = FALSE,
  seed      = MASTER_SEED
)

rsf_hp <- list(
  ntree    = 1000L,
  mtry     = as.integer(rsf_tune$optimal["mtry"]),
  nodesize = as.integer(rsf_tune$optimal["nodesize"])
)

cat(sprintf("RSF tuned (OOB): mtry=%d, nodesize=%d, ntree=%d\n",
            rsf_hp$mtry, rsf_hp$nodesize, rsf_hp$ntree))
```

# Sanity checks

**(a) Inverse-transform formula.** If sampling is correct then
$A_0\tilde T_i^2 e^{\beta^\top X_i} = -\ln U_i \sim \mathrm{Exp}(1)$, so mean
and variance should both be $\approx 1$.

```{r check-inv}
exp1_chk <- A0 * Ttilde_pilot^2 * exp(as.vector(X_pilot %*% BETA))
cat(sprintf("Exp(1) check: mean=%.3f (~1), var=%.3f (~1)\n",
            mean(exp1_chk), var(exp1_chk)))
```

**(b) Monotonicity of $S$.**

```{r check-surv}
s_chk <- as.vector(true_survival_matrix(X_pilot[1, , drop = FALSE], time_grid))
cat(sprintf("S range [%.3f, %.3f]; monotone decreasing: %s\n",
            min(s_chk), max(s_chk), all(diff(s_chk) <= 0)))
```

**(c) $\Sigma$ — empirical vs theoretical correlation.**

```{r check-sigma}
cat(sprintf("max |empirical corr - Sigma| = %.3f\n",
            max(abs(cor(X_pilot) - Sigma))))
```

**(d) Selection is present.** Truncation retains subjects who survived to their
entry time, which enriches the retained set for low $\beta^\top X$. If the
retained mean linear predictor is not below the pool mean, the filter is not
doing what the design assumes.

```{r check-selection}
lp_pool <- as.vector(X_pilot %*% BETA)
cat(sprintf("mean beta'X   pool = %+.4f | retained = %+.4f\n",
            mean(lp_pool), mean(lp_pool[kept])))
```

Note that this shift does not by itself bias the target. The estimand is the
conditional $S(t\mid x)$, and Stage 1 evaluates on covariates drawn from the
population distribution, so a shifted marginal distribution of $X$ in the
training sample is not the mechanism of interest. What matters is the risk set:
a subject enters it only at $R_i$, and under $R \perp (\tilde T, X)$ the
delayed-entry risk set at each $t$ is a random subsample of the population risk
set at $t$.

**(e) Pool headroom.** The pool must retain at least `N_LTRC` subjects.

```{r check-pool}
cat(sprintf("need %d, expect ~%.0f retained from a pool of %d\n",
            N_LTRC, (1 - TARGET_TRUNC) * N_POOL, N_POOL))
```

# Saving

```{r save}
setup <- list(
  Sigma     = Sigma,
  theta     = theta_hat,
  rsf_hp    = rsf_hp,
  time_grid = time_grid,
  beta      = BETA,
  ltrc      = list(
    cL       = C_L,
    cU       = cU_hat,
    theta_lt = theta_lt_hat,
    n_pool   = N_POOL,
    n_ltrc   = N_LTRC
  ),
  meta      = list(
    n_pilot       = N_PILOT,
    target_censor = TARGET_CENSOR,
    emp_censor    = emp_censor,
    target_trunc  = TARGET_TRUNC,
    emp_retained  = mean(kept),
    emp_censor_lt = mean(df_pilot$status == 0),
    grid_source   = "ltrc",
    tuned_on      = "ltrc",
    sigma_type    = SIGMA_TYPE,
    sigma_rho     = SIGMA_RHO,
    master_seed   = MASTER_SEED
  )
)
saveRDS(setup, "setup.rds")
cat("setup.rds saved — ready for Stage 1.\n")
```
