---
title: "Survival Simulation 05 — Stage 0: Setup (DGP, Calibration, Tuning)"
subtitle: "RIST paper Scenario 3 (Gamma survival times, high-dimensional correlated), with left truncation"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# Background

This is the fifth simulation in this project, reusing the two-stage format
established by simulations 01-04 — same shared engine
(`../_common/sim_engine.R`), same reporting (`../_common/sim_report.R`) — on
a fifth DGP.

The DGP is Scenario 3 from Zhu and Kosorok (2012), "Recursively Imputed
Survival Trees," Section 4.1
(`../../papers/Recursively Imputed Survival Trees.pdf`, p. 334): survival
times drawn from a **Gamma** distribution, not Exponential — a genuinely
different distributional family from simulations 03 and 04, not just a
different link function.

## What is taken from the paper, and what we add

Taken from RIST's Scenario 3 (unchanged):

- $p=25$, $X\sim N(0,V)$, $V_{ij}=\rho^{|i-j|}$, $\rho=0.75$ (simulation
  03's Scenario 1 uses $\rho=0.9$; this is a milder correlation).
- $\mu(X) = 0.5 + 0.3\,|\sum_{i=11}^{15}x_i|$ — only 5 of the 25 covariates
  ($x_{11},\dots,x_{15}$) drive the model; the rest are noise.
- $T\sim\mathrm{Gamma}(\text{shape}=\mu(X),\ \text{scale}=2)$
- $C\sim\mathrm{Uniform}(0,U)$, a single shared upper bound
- $\tau=10$ (administrative censoring)
- target censoring rate $\approx20\%$

## A subtlety this scenario has that simulation 04 does not

In simulation 04 (Scenario 2), the censoring window's own bound ($\tau=6$)
*was* the administrative cutoff — $C$ could never exceed it, so there was
nothing else to fold in. Here the paper specifies the censoring window as
$\mathrm{Unif}(0, 1.5\tau)$ — **wider** than $\tau$ itself. So a subject can
be censored two structurally different ways: $C$ landing before $\tilde T$
while both are still under $\tau$, or $\tilde T$ itself exceeding $\tau$
(administrative censoring, regardless of $C$). `censoring_rate()` and
`censoring_rate_lt()` in `dgp.R` branch on this explicitly, rather than
folding $\tau$ into a single expression the way simulations 02/03 do for
their memoryless Exponential censoring — that folding is only exact when
the censoring distribution is memoryless, which a bounded Uniform window is
not. The payoff shows up in Stage 0's own numbers below: analytic and
empirical censoring rates land much closer together here than in earlier
simulations.

Added here, present in neither the paper nor this scenario (Scenario 3 has
no left truncation):

- entry time $R\sim\mathrm{Unif}(c_L,c_U)$, retention filtered on
  $R\le\tilde T$ — the same mechanism simulations 01-04 use.
- residual censoring after entry, keeping the paper's own **Uniform**
  shape (as in simulation 04): $C = R + \mathrm{Unif}(0, w_{LT})$.
- $c_U$, $U$ (reference), and $w_{LT}$ calibrated by root-find against a
  target rate, following this project's established pattern, rather than
  taken directly from the paper's fixed $1.5\tau=15$. RIST's
  $\approx20\%$ is the design target for both.
- sample sizes, matching simulations 01-04 rather than the paper's own
  $n=300$: a population pool of $N_{\text{pool}} = 50{,}000$ filtered down
  to $N_{\text{ltrc}} = 5{,}000$ per replication.
- RSF hyperparameters: RIST's Section 4.2 recipe recomputed for $p=25$
  (same as simulation 03): $\lfloor\sqrt{25}\rfloor=5$, $n_{\min}=6$.

```{r libs}
source("dgp.R")
suppressPackageStartupMessages({
  library(survival)
  library(randomForestSRC)
})
```

Outputs of this stage, all written to `setup.rds`:

1. **$\Sigma$** — covariate correlation matrix (AR(1), $\rho=0.75$, PD).
2. **$U$** — right-censoring upper bound for the untruncated reference DGP,
   calibrated to $\approx20\%$ censoring.
3. **$c_U$** — entry-window upper bound, calibrated so that $\approx\tfrac13$
   of the population is truncated away.
4. **$w_{LT}$** — censoring width under truncation, calibrated to
   $\approx20\%$ censoring among the *retained* subjects.
5. **Time grid** $t_1,\dots,t_M$.
6. **RSF hyperparameters**.

## Ordering

As in simulations 01-04, the time grid and RSF tuning are derived from a
**left-truncated** pilot sample, so the sequence runs
$\Sigma \to U \to c_U \to w_{LT} \to$ pilot sample $\to$ grid $\to$ tuning.

```{r params}
MASTER_SEED   <- 20250321L   # distinct from simulations 01-04's seeds

# --- calibration targets ---
N_PILOT       <- 50000L
TARGET_CENSOR <- 0.20        # RIST Scenario 3's stated ~20%, as a design target
TARGET_TRUNC  <- 1/3         # same target as simulations 01-04
M_TIMES       <- 100L
TIME_QUANTILE <- 0.90
SIGMA_RHO     <- 0.75        # RIST Scenario 3's value (Scenario 1 uses 0.9)
SIGMA_TYPE    <- "ar1"

# --- left-truncation settings ---
N_POOL        <- 50000L
N_LTRC        <- 5000L       # matches simulations 01-04's run size
C_L           <- 0

# --- RSF tuning ---
# See simulation 01's Stage 0 for why RUN_TUNE defaults off. RUN_TUNE=FALSE
# falls back to RIST's own RSF settings (Section 4.2), recomputed for
# p=25 (same value as simulation 03, which also has p=25).
RUN_TUNE      <- FALSE
RSF_MTRY_DEFAULT     <- as.integer(floor(sqrt(P)))   # RIST Sec. 4.2: K = floor(sqrt(p))
RSF_NODESIZE_DEFAULT <- 6L                            # RIST Sec. 4.2: n_min = 6
```

# Correlation matrix $\Sigma$

```{r sigma}
Sigma <- build_sigma(P, SIGMA_RHO, SIGMA_TYPE)
stopifnot(min(eigen(Sigma, symmetric = TRUE, only.values = TRUE)$values) > 0)

cat(sprintf("Sigma %dx%d (%s, rho=%.2f) — PD OK\n",
            nrow(Sigma), ncol(Sigma), SIGMA_TYPE, SIGMA_RHO))
round(Sigma[1:8, 1:8], 3)   # 25x25 is large to print in full
```

# Censoring upper bound $U$ (untruncated reference)

$C \perp \tilde T$, so the censoring rate integrates out in closed form
given the pilot's $\tilde T$ values, branching on whether $\tilde T$
exceeds the administrative cutoff (see the note above):
$$g(U)=\frac1n\sum_i \begin{cases} 1 & \tilde T_i > \tau \\ \min(\tilde T_i, U)/U & \tilde T_i \le \tau \end{cases}$$
which is monotone decreasing in $U$ on the branch where it matters.
`uniroot` therefore returns a unique root with no sampling noise.

This parameter belongs to the untruncated DGP. Stage 1 fits only truncated
data, so it is carried in `setup.rds` for reference rather than consumed
downstream.

```{r censoring-U}
set.seed(MASTER_SEED)

X_pilot      <- sample_X(N_PILOT, Sigma)
Ttilde_pilot <- sample_event_times(X_pilot)

U_hat <- uniroot(
  f        = function(u) censoring_rate(u, Ttilde_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 200),
  tol      = 1e-6
)$root

chk_df     <- make_survival_sample(N_PILOT, Sigma, U_hat)
emp_censor <- mean(chk_df$status == 0)

cat(sprintf("U = %.4f  (paper's own 1.5*tau = 15, for comparison)\n", U_hat))
cat(sprintf("  censoring  analytic = %.3f | empirical = %.3f\n",
            censoring_rate(U_hat, Ttilde_pilot), emp_censor))
cat(sprintf("  P(T~ > tau=%d) = %.4f  (handled exactly, not folded/approximated)\n",
            TAU, mean(Ttilde_pilot > TAU)))
```

# Entry window $c_U$

Integrating out $R \sim \mathrm{Unif}(c_L, c_U)$ given the pilot's $\tilde T$:
$$\mathbb P(R > \tilde T_i) =
\begin{cases}
1 & \tilde T_i < c_L\\[2pt]
\dfrac{c_U - \tilde T_i}{c_U - c_L} & c_L \le \tilde T_i \le c_U\\[6pt]
0 & \tilde T_i > c_U
\end{cases}$$
increasing in $c_U$.

```{r cU}
cU_hat <- uniroot(
  f        = function(cu) trunc_rate(cu, Ttilde_pilot, C_L) - TARGET_TRUNC,
  interval = c(C_L + 1e-6, 50 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("cU = %.4f  (entry window: Unif(%.2f, %.2f))\n", cU_hat, C_L, cU_hat))
cat(sprintf("  truncated away  analytic = %.3f  (target %.3f)\n",
            trunc_rate(cU_hat, Ttilde_pilot, C_L), TARGET_TRUNC))
```

# Censoring under truncation, $w_{LT}$

Calibrated on the **retained** subjects, since that is the population the
models see. Same branching as $U$ above, with $d_i = \tilde T_i - R_i$ in
place of $\tilde T_i$ on the non-administrative branch.

```{r w-lt}
set.seed(MASTER_SEED + 2L)

R_pilot  <- runif(N_PILOT, C_L, cU_hat)
kept     <- R_pilot <= Ttilde_pilot
Tk_pilot <- Ttilde_pilot[kept]
Rk_pilot <- R_pilot[kept]

w_lt_hat <- uniroot(
  f        = function(w) censoring_rate_lt(w, Tk_pilot, Rk_pilot) - TARGET_CENSOR,
  interval = c(1e-6, 200),
  tol      = 1e-6
)$root

cat(sprintf("retained %d / %d pool subjects (%.1f%%)\n",
            sum(kept), N_PILOT, 100 * mean(kept)))
cat(sprintf("w_LT = %.4f  (paper's own 1.5*tau = 15, for comparison)\n", w_lt_hat))
cat(sprintf("  censoring (retained)  analytic = %.3f\n",
            censoring_rate_lt(w_lt_hat, Tk_pilot, Rk_pilot)))
```

Both $U$ and $w_{LT}$ landing close to the paper's own $1.5\tau=15$ is a
useful check on the calibration.

# The left-truncated pilot sample

One sample at the Stage 1 run size, used for both the grid and the tuning.

```{r ltrc-pilot}
set.seed(MASTER_SEED + 3L)
df_pilot <- make_ltrc_sample(N_LTRC, N_POOL, Sigma, w_lt_hat, cU_hat, C_L)

cat(sprintf("LTRC pilot: n=%d,  censoring=%.3f,  entry <= time on all rows: %s\n",
            nrow(df_pilot), mean(df_pilot$status == 0),
            all(df_pilot$entry <= df_pilot$time)))
```

# Time grid

As in simulations 01-04, taken from the **left-truncated** pilot — the
regime Stage 1 fits — up to the 90th percentile of observed follow-up.

```{r grid}
upper     <- as.numeric(quantile(df_pilot$time, TIME_QUANTILE))
time_grid <- seq(0, upper, length.out = M_TIMES + 1L)[-1L]

cat(sprintf("time grid: M=%d, range (%.3f, %.3f]\n",
            length(time_grid), min(time_grid), max(time_grid)))
cat(sprintf("cU = %.3f sits at grid position %d of %d\n",
            cU_hat, sum(time_grid < cU_hat) + 1L, length(time_grid)))
```

# RSF hyperparameters

Same choices as simulations 01-04 when tuning does run: on the
left-truncated pilot, and entry-agnostic. `RUN_TUNE` gates the search; with
`RUN_TUNE <- FALSE`, `rsf_hp` falls back to RIST's Section 4.2 recipe for
$p=25$: `mtry = floor(sqrt(25)) = 5`, `nodesize = 6`.

```{r rsf-tune, eval=RUN_TUNE}
rsf_tune <- tune(
  Surv(time, status) ~ .,
  data      = df_pilot[, c(grep("^x", names(df_pilot), value = TRUE),
                           "time", "status")],
  ntree.try = 300,
  do.best   = FALSE,
  seed      = MASTER_SEED
)
```

```{r rsf-hp}
rsf_hp <- if (RUN_TUNE) {
  list(ntree    = 1000L,
       mtry     = as.integer(rsf_tune$optimal["mtry"]),
       nodesize = as.integer(rsf_tune$optimal["nodesize"]))
} else {
  list(ntree = 500L, mtry = RSF_MTRY_DEFAULT, nodesize = RSF_NODESIZE_DEFAULT)
}

cat(sprintf("RSF hyperparameters (%s): mtry=%d, nodesize=%d, ntree=%d\n",
            if (RUN_TUNE) "OOB-tuned" else "fixed, RIST Sec. 4.2 recipe",
            rsf_hp$mtry, rsf_hp$nodesize, rsf_hp$ntree))
```

# Sanity checks

**(a) $\mu(X)$ range.** Should stay $\ge 0.5$ (the intercept), since the
slope term is non-negative by construction ($0.3\times|\cdot|$).

```{r check-mu}
mu_chk <- mu_shape(X_pilot)
cat(sprintf("mu range [%.4f, %.4f], any below 0.5: %s\n",
            min(mu_chk), max(mu_chk), any(mu_chk < MU_INTERCEPT - 1e-8)))
```

**(b) Monotonicity of $S$.**

```{r check-surv}
s_chk <- as.vector(true_survival_matrix(X_pilot[1, , drop = FALSE], time_grid))
cat(sprintf("S range [%.3f, %.3f]; monotone decreasing: %s\n",
            min(s_chk), max(s_chk), all(diff(s_chk) <= 0)))
```

**(c) $\Sigma$ — empirical vs theoretical correlation.**

```{r check-sigma}
cat(sprintf("max |empirical corr - Sigma| = %.3f\n",
            max(abs(cor(X_pilot) - Sigma))))
```

**(d) Selection is present.** As in simulations 03/04 (and unlike
simulation 01), larger $\mu(X)$ means a larger Gamma mean
($\text{mean}=\mu(X)\times\text{scale}$), i.e. longer $\tilde T$ on
average, so the retained set should skew toward **larger** $\mu(X)$.

```{r check-selection}
mu_pool <- mu_shape(X_pilot)
cat(sprintf("mean mu(X)   pool = %.4f | retained = %.4f\n",
            mean(mu_pool), mean(mu_pool[kept])))
```

Note that this shift does not by itself bias the target. The estimand is
the conditional $S(t\mid x)$, and Stage 1 evaluates on covariates drawn
from the population distribution. What matters is the risk set: under
$R \perp (\tilde T, X)$ the delayed-entry risk set at each $t$ is a random
subsample of the population risk set at $t$.

**(e) Pool headroom.** The pool must retain at least `N_LTRC` subjects.

```{r check-pool}
cat(sprintf("need %d, expect ~%.0f retained from a pool of %d\n",
            N_LTRC, (1 - TARGET_TRUNC) * N_POOL, N_POOL))
```

# Saving

```{r save}
setup <- list(
  Sigma     = Sigma,
  rsf_hp    = rsf_hp,
  time_grid = time_grid,
  cens_U    = U_hat,
  tau       = TAU,
  ltrc      = list(
    cL     = C_L,
    cU     = cU_hat,
    w_lt   = w_lt_hat,
    n_pool = N_POOL,
    n_ltrc = N_LTRC
  ),
  meta      = list(
    n_pilot       = N_PILOT,
    target_censor = TARGET_CENSOR,
    emp_censor    = emp_censor,
    target_trunc  = TARGET_TRUNC,
    emp_retained  = mean(kept),
    emp_censor_lt = mean(df_pilot$status == 0),
    grid_source   = "ltrc",
    tuned_on      = if (RUN_TUNE) "ltrc" else "fixed (RIST Sec. 4.2 recipe, RUN_TUNE=FALSE)",
    sigma_type    = SIGMA_TYPE,
    sigma_rho     = SIGMA_RHO,
    master_seed   = MASTER_SEED,
    source        = "Recursively Imputed Survival Trees.pdf, Scenario 3 (p.334)"
  )
)
saveRDS(setup, "setup.rds")
cat("setup.rds saved — ready for Stage 1.\n")
```
