---
title: "Survival Simulation 03 — Stage 0: Setup (DGP, Calibration, Tuning)"
subtitle: "RIST paper Scenario 1 (high-dimensional, correlated Cox PH), with left truncation"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# Background

This is the third simulation in this project, reusing the two-stage format
established by simulations 01 and 02 — same shared engine
(`../_common/sim_engine.R`), same reporting (`../_common/sim_report.R`) — on
a third DGP.

The DGP is Scenario 1 from Zhu and Kosorok (2012), "Recursively Imputed
Survival Trees," Section 4.1
(`../../papers/Recursively Imputed Survival Trees.pdf`, p. 334): a
proportional-hazards model with 25 correlated covariates. RIST states that
Scenario 1 is itself "adapted from Section 4 of" Ishwaran, Kogalur,
Gorodeski, Minn, and Lauer (2010), "High-Dimensional Variable Selection for
Survival Data,"
(`../../papers/High-Dimensional_Variable_Selection_for_Survival_D.pdf`) —
see the note below on why that source paper mattered here.

## A correction, checked against the source paper

RIST's Scenario 1 gives the survival-time mean as
$$\mu = b_0 \times \textstyle\sum_{i=11}^{20} x_i,$$
a raw linear sum. Since $X$ is standard normal, this sum is mean-zero and
therefore negative about half the time — not a valid mean for an Exponential
distribution. The paper it cites resolves this: Ishwaran et al.'s own
Section 4 defines $\mu = \exp\!\big(\sum_k \beta_k x_k\big)$ — the sum is a
linear predictor that gets **exponentiated**, not the mean itself. We follow
that exp() form here; RIST's restatement is read as having dropped it in
transcription. `dgp.R` documents this at the top.

Two further changes RIST *does* make deliberately relative to its source
(kept as given, since these are not the ambiguous part): the active index
range widens from 5 middle coordinates to 10 ($x_{11},\dots,x_{20}$), with a
single fixed $b_0=0.1$ rather than the source's five tested values; and the
censoring mean is **half** the average of $\mu$, rather than the source's
plain average.

## What is taken from the papers, and what we add

Taken from RIST's Scenario 1 (unchanged):

- $p=25$, $X\sim N(0,V)$, $V_{ij}=\rho^{|i-j|}$, $\rho=0.9$
- $\mu(X)=\exp(b_0\sum_{i=11}^{20}x_i)$, $b_0=0.1$ (exp() restored per above)
- $T\sim\mathrm{Exponential}(\text{mean}=\mu(X))$
- $C\sim\mathrm{Exponential}(\text{mean}=\theta)$, a single shared $\theta$
- $\tau=4$ (administrative censoring)
- target censoring rate $\approx 30\%$

Added here, present in neither paper (Scenario 1 has no left truncation):

- entry time $R\sim\mathrm{Unif}(c_L,c_U)$, retention filtered on
  $R\le\tilde T$, censoring residual after entry
  ($C=R+\mathrm{Exp}(\theta_{LT})$) so `entry <= time` holds on every row —
  the same mechanism simulations 01 and 02 use.
- $c_U$, $\theta$, and $\theta_{LT}$ calibrated by root-find against a
  target rate, following this project's established pattern, rather than
  computed once from a closed-form sample average as RIST does. RIST's
  $\approx30\%$ is treated as the design target for **both** the
  untruncated reference $\theta$ and the retained-population $\theta_{LT}$
  — the same resolution simulation 02 used for its own paper-censoring
  discrepancy.
- sample sizes: RIST's Scenario 1 uses $n=200$ per replication (it is
  deliberately a small-$n$, high-$p$ stress test for variable selection).
  Here, matching simulations 01/02, a population pool of $N_{\text{pool}} =
  50{,}000$ is filtered down to a retained training set of $N_{\text{ltrc}}
  = 5{,}000$ per replication.
- RSF hyperparameters: `mtry` and `nodesize`, when not OOB-tuned (see
  below), fall back to the RIST paper's own tuning settings (Section 4.2:
  $K=\lfloor\sqrt p\rfloor$, $n_{\min}=6$) rather than an arbitrary value —
  $\lfloor\sqrt{25}\rfloor=5$.

```{r libs}
source("dgp.R")
suppressPackageStartupMessages({
  library(survival)
  library(randomForestSRC)
})
```

Outputs of this stage, all written to `setup.rds`:

1. **$\Sigma$** — covariate correlation matrix (AR(1), $\rho=0.9$, PD).
2. **$\theta$** — right-censoring parameter for the untruncated reference DGP,
   calibrated to $\approx30\%$ censoring.
3. **$c_U$** — entry-window upper bound, calibrated so that $\approx\tfrac13$
   of the population is truncated away (same target as simulations 01/02,
   added here rather than in either paper).
4. **$\theta_{LT}$** — censoring parameter under truncation, calibrated to
   $\approx30\%$ censoring among the *retained* subjects.
5. **Time grid** $t_1,\dots,t_M$.
6. **RSF hyperparameters** — tuned by out-of-bag error, or the paper's own
   fixed recipe when tuning is off.

## Ordering

As in simulations 01/02, the time grid and RSF tuning are derived from a
**left-truncated** pilot sample, the regime Stage 1 fits, so the sequence
runs $\Sigma \to \theta \to c_U \to \theta_{LT} \to$ pilot sample $\to$ grid
$\to$ tuning.

```{r params}
MASTER_SEED   <- 20250107L   # distinct from simulations 01 and 02's seeds

# --- calibration targets ---
N_PILOT       <- 50000L
TARGET_CENSOR <- 0.30        # RIST Scenario 1's stated ~30%, as a design target
TARGET_TRUNC  <- 1/3         # same target as simulations 01/02
M_TIMES       <- 100L
TIME_QUANTILE <- 0.90
SIGMA_RHO     <- 0.9         # RIST Scenario 1's value (01 uses 0.2)
SIGMA_TYPE    <- "ar1"

# --- left-truncation settings ---
N_POOL        <- 50000L
N_LTRC        <- 5000L       # matches simulations 01/02's run size
C_L           <- 0

# --- RSF tuning ---
# See simulation 01's Stage 0 for why RUN_TUNE defaults off: tuning on the
# left-truncated pilot tends to select a small nodesize, producing deep,
# memory-heavy trees. With p=25 here (vs. 5-6 in 01/02) the risk is at least
# as large. RUN_TUNE=FALSE falls back to RIST's own RSF settings (Section
# 4.2) rather than an arbitrary previous value: K = floor(sqrt(p)) = 5,
# nmin = 6. Flip RUN_TUNE once running on a machine with more memory.
RUN_TUNE      <- FALSE
RSF_MTRY_DEFAULT     <- as.integer(floor(sqrt(P)))   # RIST Sec. 4.2: K = floor(sqrt(p))
RSF_NODESIZE_DEFAULT <- 6L                            # RIST Sec. 4.2: n_min = 6
```

# Correlation matrix $\Sigma$

```{r sigma}
Sigma <- build_sigma(P, SIGMA_RHO, SIGMA_TYPE)
stopifnot(min(eigen(Sigma, symmetric = TRUE, only.values = TRUE)$values) > 0)

cat(sprintf("Sigma %dx%d (%s, rho=%.2f) — PD OK\n",
            nrow(Sigma), ncol(Sigma), SIGMA_TYPE, SIGMA_RHO))
round(Sigma[1:8, 1:8], 3)   # 25x25 is large to print in full
```

# Censoring parameter $\theta$ (untruncated reference)

Because $C \perp \tilde T$, the censoring rate integrates out in closed form
given the pilot's $\tilde T$ values (with admin censoring at $\tau$ folded
in),
$$g(\theta)=\frac1n\sum_i\Big(1-e^{-\min(\tilde T_i,\tau)/\theta}\Big),$$
which is monotone decreasing in $\theta$. `uniroot` therefore returns a
unique root with no sampling noise.

This parameter belongs to the untruncated DGP. Stage 1 fits only truncated
data, so it is carried in `setup.rds` for reference rather than consumed
downstream.

```{r theta}
set.seed(MASTER_SEED)

X_pilot      <- sample_X(N_PILOT, Sigma)
Ttilde_pilot <- sample_event_times(X_pilot)

theta_hat <- uniroot(
  f        = function(th) censoring_rate(th, Ttilde_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 10 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

chk_df     <- make_survival_sample(N_PILOT, Sigma, theta_hat)
emp_censor <- mean(chk_df$status == 0)

cat(sprintf("theta = %.4f\n  censoring  analytic = %.3f | empirical = %.3f\n",
            theta_hat, censoring_rate(theta_hat, Ttilde_pilot), emp_censor))
cat(sprintf("  P(T~ > tau=%d) = %.4f  (administrative censoring is a minor contributor)\n",
            TAU, mean(Ttilde_pilot > TAU)))
```

# Entry window $c_U$

Integrating out $R \sim \mathrm{Unif}(c_L, c_U)$ given the pilot's $\tilde T$:
$$\mathbb P(R > \tilde T_i) =
\begin{cases}
1 & \tilde T_i < c_L\\[2pt]
\dfrac{c_U - \tilde T_i}{c_U - c_L} & c_L \le \tilde T_i \le c_U\\[6pt]
0 & \tilde T_i > c_U
\end{cases}$$
increasing in $c_U$.

```{r cU}
cU_hat <- uniroot(
  f        = function(cu) trunc_rate(cu, Ttilde_pilot, C_L) - TARGET_TRUNC,
  interval = c(C_L + 1e-6, 50 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("cU = %.4f  (entry window: Unif(%.2f, %.2f);  cU / tau = %.3f)\n",
            cU_hat, C_L, cU_hat, cU_hat / TAU))
cat(sprintf("  truncated away  analytic = %.3f  (target %.3f)\n",
            trunc_rate(cU_hat, Ttilde_pilot, C_L), TARGET_TRUNC))
```

# Censoring under truncation, $\theta_{LT}$

Calibrated on the **retained** subjects, since that is the population the
models see. With $C_i = R_i + E_i$ and $E_i\sim\mathrm{Exp}(\text{mean}=\theta_{LT})$,
$$\mathbb P(C_i < \tilde T_i)=\mathbb P(E_i < \min(\tilde T_i,\tau) - R_i)
= 1 - e^{-(\min(\tilde T_i,\tau) - R_i)/\theta_{LT}},$$
decreasing in $\theta_{LT}$.

```{r theta-lt}
set.seed(MASTER_SEED + 2L)

R_pilot  <- runif(N_PILOT, C_L, cU_hat)
kept     <- R_pilot <= Ttilde_pilot
Tk_pilot <- Ttilde_pilot[kept]
Rk_pilot <- R_pilot[kept]

theta_lt_hat <- uniroot(
  f        = function(th) censoring_rate_lt(th, Tk_pilot, Rk_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 50 * max(Tk_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("retained %d / %d pool subjects (%.1f%%)\n",
            sum(kept), N_PILOT, 100 * mean(kept)))
cat(sprintf("theta_LT = %.4f\n  censoring (retained)  analytic = %.3f\n",
            theta_lt_hat, censoring_rate_lt(theta_lt_hat, Tk_pilot, Rk_pilot)))
```

# The left-truncated pilot sample

One sample at the Stage 1 run size, used for both the grid and the tuning.

```{r ltrc-pilot}
set.seed(MASTER_SEED + 3L)
df_pilot <- make_ltrc_sample(N_LTRC, N_POOL, Sigma, theta_lt_hat, cU_hat, C_L)

cat(sprintf("LTRC pilot: n=%d,  censoring=%.3f,  entry <= time on all rows: %s\n",
            nrow(df_pilot), mean(df_pilot$status == 0),
            all(df_pilot$entry <= df_pilot$time)))
```

# Time grid

As in simulations 01/02, taken from the **left-truncated** pilot — the
regime Stage 1 fits — up to the 90th percentile of observed follow-up.

```{r grid}
upper     <- as.numeric(quantile(df_pilot$time, TIME_QUANTILE))
time_grid <- seq(0, upper, length.out = M_TIMES + 1L)[-1L]

cat(sprintf("time grid: M=%d, range (%.3f, %.3f]\n",
            length(time_grid), min(time_grid), max(time_grid)))
cat(sprintf("cU = %.3f sits at grid position %d of %d\n",
            cU_hat, sum(time_grid < cU_hat) + 1L, length(time_grid)))
```

# RSF hyperparameters

`tune()` scans $(\texttt{nodesize}, \texttt{mtry})$ and minimises OOB error.
Same two choices as simulations 01/02 when tuning does run: on the
left-truncated pilot (the regime Stage 1 fits), and entry-agnostic (`tune()`
carves out an internal holdout without subsetting extra arguments alongside
it, and hyperparameters should not be selected using information only two
of the four Stage 1 arms receive).

`RUN_TUNE` gates the search (`eval=RUN_TUNE`, so the chunk is skipped
entirely). With `RUN_TUNE <- FALSE`, `rsf_hp` falls back to RIST's own
Section 4.2 recipe — `mtry = floor(sqrt(p)) = 5`, `nodesize = 6` — rather
than a value carried over from a previous run, since there is no prior run
of this DGP to carry over from.

```{r rsf-tune, eval=RUN_TUNE}
rsf_tune <- tune(
  Surv(time, status) ~ .,
  data      = df_pilot[, c(grep("^x", names(df_pilot), value = TRUE),
                           "time", "status")],
  ntree.try = 300,
  do.best   = FALSE,
  seed      = MASTER_SEED
)
```

```{r rsf-hp}
rsf_hp <- if (RUN_TUNE) {
  list(ntree    = 1000L,
       mtry     = as.integer(rsf_tune$optimal["mtry"]),
       nodesize = as.integer(rsf_tune$optimal["nodesize"]))
} else {
  list(ntree = 500L, mtry = RSF_MTRY_DEFAULT, nodesize = RSF_NODESIZE_DEFAULT)
}

cat(sprintf("RSF hyperparameters (%s): mtry=%d, nodesize=%d, ntree=%d\n",
            if (RUN_TUNE) "OOB-tuned" else "fixed, RIST Sec. 4.2 recipe",
            rsf_hp$mtry, rsf_hp$nodesize, rsf_hp$ntree))
```

# Sanity checks

**(a) Inverse-transform formula.** If sampling is correct then
$\tilde T_i \exp(-\beta^\top X_i) = -\ln U_i \sim \mathrm{Exp}(1)$, so mean
and variance should both be $\approx 1$.

```{r check-inv}
exp1_chk <- Ttilde_pilot * exp(-as.vector(X_pilot %*% BETA))
cat(sprintf("Exp(1) check: mean=%.3f (~1), var=%.3f (~1)\n",
            mean(exp1_chk), var(exp1_chk)))
```

**(b) Monotonicity of $S$.**

```{r check-surv}
s_chk <- as.vector(true_survival_matrix(X_pilot[1, , drop = FALSE], time_grid))
cat(sprintf("S range [%.3f, %.3f]; monotone decreasing: %s\n",
            min(s_chk), max(s_chk), all(diff(s_chk) <= 0)))
```

**(c) $\Sigma$ — empirical vs theoretical correlation.**

```{r check-sigma}
cat(sprintf("max |empirical corr - Sigma| = %.3f\n",
            max(abs(cor(X_pilot) - Sigma))))
```

**(d) Selection is present, in the opposite direction from simulation 01.**
In simulation 01, higher $\beta^\top X$ raises the *hazard*, so it shortens
$\tilde T$ and those subjects are disproportionately truncated away — the
retained set skews toward low $\beta^\top X$. Here $\mu(X)=\exp(\beta^\top X)$
is the *mean* of $\tilde T$ directly, so higher $\beta^\top X$ means **longer**
$\tilde T$ on average, and it is low-$\beta^\top X$ subjects (shorter $\tilde
T$) that get filtered out before their entry time. The retained set should
therefore skew toward **high** $\beta^\top X$ — the opposite sign from
simulation 01's check.

```{r check-selection}
lp_pool <- as.vector(X_pilot %*% BETA)
cat(sprintf("mean beta'X   pool = %+.4f | retained = %+.4f\n",
            mean(lp_pool), mean(lp_pool[kept])))
```

Note that this shift does not by itself bias the target. The estimand is the
conditional $S(t\mid x)$, and Stage 1 evaluates on covariates drawn from the
population distribution, so a shifted marginal distribution of $X$ in the
training sample is not the mechanism of interest. What matters is the risk
set: a subject enters it only at $R_i$, and under $R \perp (\tilde T, X)$ the
delayed-entry risk set at each $t$ is a random subsample of the population
risk set at $t$.

**(e) Pool headroom.** The pool must retain at least `N_LTRC` subjects.

```{r check-pool}
cat(sprintf("need %d, expect ~%.0f retained from a pool of %d\n",
            N_LTRC, (1 - TARGET_TRUNC) * N_POOL, N_POOL))
```

# Saving

```{r save}
setup <- list(
  Sigma     = Sigma,
  theta     = theta_hat,
  rsf_hp    = rsf_hp,
  time_grid = time_grid,
  beta      = BETA,
  tau       = TAU,
  ltrc      = list(
    cL       = C_L,
    cU       = cU_hat,
    theta_lt = theta_lt_hat,
    n_pool   = N_POOL,
    n_ltrc   = N_LTRC
  ),
  meta      = list(
    n_pilot       = N_PILOT,
    target_censor = TARGET_CENSOR,
    emp_censor    = emp_censor,
    target_trunc  = TARGET_TRUNC,
    emp_retained  = mean(kept),
    emp_censor_lt = mean(df_pilot$status == 0),
    grid_source   = "ltrc",
    tuned_on      = if (RUN_TUNE) "ltrc" else "fixed (RIST Sec. 4.2 recipe, RUN_TUNE=FALSE)",
    sigma_type    = SIGMA_TYPE,
    sigma_rho     = SIGMA_RHO,
    master_seed   = MASTER_SEED,
    source        = "Recursively Imputed Survival Trees.pdf, Scenario 1 (p.334); mu formula per High-Dimensional_Variable_Selection_for_Survival_D.pdf, Sec.4"
  )
)
saveRDS(setup, "setup.rds")
cat("setup.rds saved — ready for Stage 1.\n")
```
