---
title: "Survival Simulation 06 — Stage 1: Entry Times on RIST Scenario 4"
subtitle: "One sample, two phases: entry zeroed, then entry supplied"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# The experiment

Same design as simulations 01-05's Stage 1, run against a sixth and final
DGP. Every replication draws **one** left-truncated sample and runs it
through **two phases**. The rows, the seed, and the hyperparameters are
identical throughout; the only thing that changes between phases is the
entry vector handed to the entry-aware arms.

| | phase 1 — entry zeroed | phase 2 — entry supplied |
|---|---|---|
| **Cox, standard** | `Surv(time, status)` | unchanged |
| **Cox, left-truncation** | `Surv(0, time, status)` | `Surv(entry, time, status)` |
| **RSF, standard** | `rfsrc(...)` | unchanged |
| **RSF, left-truncation** | `rfsrc(..., entry.time = 0)` | `rfsrc(..., entry.time = entry)` |

The standard arms do not read the entry vector at all, so they are fitted
once and carried into both phases unchanged.

## What is different from simulations 01-05

The DGP is Scenario 4 from the RIST paper's simulation study
(`../../papers/Recursively Imputed Survival Trees.pdf`, Section 4.1,
p. 334), not reproduced here — see `dgp.R` and this simulation's Stage 0
for the model, the closed-form truth, and a structural finding about the
retained-population censoring target (the admin cutoff at $\tau=4$ alone
already exceeds RIST's stated $\approx32\%$ among retained subjects, so
the extra censoring window is disabled there — see Stage 0). Survival
times are Lognormal,
$$T\sim\mathrm{Lognormal}(\text{meanlog}=\mu(X),\ \text{sdlog}=1), \qquad \mu(X)=0.1\,|{\textstyle\sum_{i=1}^{5}}x_i|+0.1\,|{\textstyle\sum_{i=21}^{25}}x_i|,$$
a third distributional family after Exponential (simulations 03/04) and
Gamma (simulation 05). RIST calls this scenario "conditionally independent
censoring": $C$ is drawn independently of $T$, but both depend on the same
$X$ through a shared location function — different from simulations 01-05,
where $C$'s distribution never depended on $X$ at all.

Left truncation is not part of the RIST paper. It is added here as in
simulations 01-05: entry times $R\sim\mathrm{Unif}(c_L,c_U)$ independent of
$(\tilde T,X)$, retention filtered on the true event time, and residual
censoring after entry — Lognormal-shaped and still $X$-dependent — so that
`entry <= time` holds on every row.

## What each phase is for

**Phase 1** should produce four equal numbers, for the same reason as
simulations 01-05: the counting-process Cox with entry 0 has the same risk
sets as the standard fit, and `entry.time = 0` is documented to reproduce
the standard forest bit for bit.

**Phase 2** supplies the observed entry times. The reported quantity is the
within-arm shift from phase 1 to phase 2 on identical data.

## Scoring

All arms are scored against the closed-form truth
$$S(t\mid X) = 1 - F_{\mathrm{Lognormal}}(t;\ \text{meanlog}=\mu(X),\ \text{sdlog}=1),$$
which truncation leaves unchanged. Test covariates are drawn from the
**population** distribution, since the estimand is the conditional
$S(t\mid x)$.

# Setup

This chunk is deliberately **not** cached — caching it would restore the
objects it creates while silently skipping its `library()` calls.

```{r libs, cache=FALSE}
source("../_common/sim_engine.R")
source("../_common/sim_report.R")
source("dgp.R")

stopifnot("entry.time" %in% names(formals(rfsrc)))
cat(sprintf("randomForestSRC %s loaded from %s\n",
            packageVersion("randomForestSRC"), path.package("randomForestSRC")))
```

```{r install, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("Shaked06/randomForestSRC@ltrc-rsf")
```

```{r load-setup, cache=FALSE}
setup <- readRDS("setup.rds")
stopifnot(!is.null(setup$ltrc))

M_TEST <- 200L      # test subjects per replication
N_REP  <- 10L       # set lower for a fast pass

str(setup[c("rsf_hp", "ltrc")])
cat(sprintf("\ntime grid: M=%d, range (%.3f, %.3f]\n",
            length(setup$time_grid), min(setup$time_grid), max(setup$time_grid)))
```

# The replication loop

Same mechanics as simulations 01-05: replications are independent, each
seeds itself from `rep_id`, and workers source the engine and `dgp.R`
rather than receiving exported closures.

`N_REP` and `M_TEST` are named in `cache.extra` so a cached loop cannot
silently survive a change to either.

```{r loop, cache.extra = list(N_REP, M_TEST, tools::md5sum(c("setup.rds", "dgp.R", "../_common/sim_engine.R")))}
res <- run_experiment(
  n_rep   = N_REP,
  setup   = setup,
  sim     = sim,
  sources = c("../_common/sim_engine.R", "dgp.R"),
  m_test  = M_TEST
)

saveRDS(res, "results_phases.rds")
```

# Results

## Phase 1 — are the left-truncation arms inert?

```{r noop, cache=FALSE}
noop_table(res)
```

The reporting chunks below are uncached: they cost milliseconds, and caching
them would let a wording change in `sim_report.R` leave stale labels on
screen.

## The two phases

```{r phase-table, cache=FALSE}
phase_table(res)
```

## Within-arm shift

```{r paired, cache=FALSE}
paired_shift(res)
```

`n_moved_down` counts replications in which the phase-2 score is the smaller
of the two.

```{r plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res)
```

## Across the time axis

```{r plot-time, cache=FALSE, fig.width=8, fig.height=5}
invisible(plot_score_by_time(res))
```

# What we have

`results_phases.rds` holds, for `r N_REP` replications, the score for eight
arm-phase combinations, the score-versus-time curve for each, and the
phase-1 no-op diagnostics — same structure as simulations 01-05, on the
last of the six DGPs in this project (the original Cox/flexihaz pair, plus
RIST Scenarios 1-4). Across all six, the entry-time correction has now been
checked under: exactly-Cox and PH-violating hazards, Exponential/Gamma/
Lognormal event times, Exponential/Uniform/X-dependent-Lognormal censoring,
and covariate dimensions from 5 to 25 with correlation from 0 to 0.9.
