---
title: "Results at a glance"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
source("_common/sim_report.R")
```

Thin summary of all six simulations' knitted results. For the DGP math, true
`S`/hazard, and PH verdicts, see `README.md` — this file is just the numbers,
computed directly from each sim's `results_phases.rds` (nothing here is
hand-typed).

## What "phase 1 → phase 2" means

Every simulation fits the *same* entry-time-aware model (`cox_lt`/`rsf_lt`) on
the *same* replication twice:

- **Phase 1**: that model is given an entry vector of all zeros — i.e.
  "pretend nothing is truncated."
- **Phase 2**: that same model is given the real, observed entry times.

So every number below is a **paired before/after comparison on one model**,
not a comparison against a different (standard) model. It happens that phase
1 (entry = 0) is mathematically proven identical to the plain/standard
model's own score (the "noop" check each sim's report also shows), so in
practice this also tells you standard-vs-entry-aware — but the actual
quantity being computed is: *does giving this model the real entry time
improve it, relative to giving it no truncation information at all?*

```{r load}
sims <- c(
  "01" = "01_cox_random_forerst_lt",
  "02" = "02_flexihaz_gompertz_lt",
  "03" = "03_rist_scenario1_lt",
  "04" = "04_rist_scenario2_lt",
  "05" = "05_rist_scenario3_lt",
  "06" = "06_rist_scenario4_lt"
)
dgp_label <- c(
  "01" = "Cox PH (exactly Cox)",
  "02" = "Gompertz-shape (PH violation)",
  "03" = "RIST Sc.1 (exactly Cox)",
  "04" = "RIST Sc.2 (nonlinear link)",
  "05" = "RIST Sc.3 (Gamma, PH violation)",
  "06" = "RIST Sc.4 (Lognormal, PH violation)"
)
metrics <- c(overall = "MAD", cindex = "C-idx", brier = "Brier")

res_list <- lapply(sims, function(d) {
  readRDS(file.path(d, "results_phases.rds"))
})
names(res_list) <- names(sims)

## % improvement from phase 1 -> phase 2, lower-is-better for every metric
## here, so positive = entry time helped.
pct_improve <- function(res, field, family) {
  pt  <- phase_table(res, field = field)
  row <- pt[pt$variant == "left-truncation" & pt$family == family, ]
  (row$phase1 - row$phase2) / row$phase1 * 100
}

summary_tbl <- do.call(rbind, lapply(names(sims), function(nm) {
  res <- res_list[[nm]]
  row <- c(`#` = nm, DGP = dgp_label[[nm]])
  for (f in names(metrics)) {
    row[paste0(metrics[[f]], " Cox")] <-
      sprintf("%.2f%%", pct_improve(res, f, "Cox"))
    row[paste0(metrics[[f]], " RSF")] <-
      sprintf("%.2f%%", pct_improve(res, f, "RSF"))
  }
  row
}))

## Read the replication count back off the saved results rather than
## hardcoding it, so this line can never go stale against the numbers it
## describes -- same rule as everything else in this file.
n_rep_all  <- vapply(res_list, function(r) as.integer(r$n_rep),  integer(1))
m_test_all <- vapply(res_list, function(r) as.integer(r$m_test), integer(1))
fmt_range <- function(v) if (length(unique(v)) == 1) as.character(v[1]) else
  paste0(min(v), "-", max(v), " (varies by sim)")
```

> **Run size:** `r fmt_range(n_rep_all)` replications per simulation,
> `r fmt_range(m_test_all)` test subjects each.
> `r if (max(n_rep_all) < 100) "This is a smoke-test run — enough to confirm the pipeline works end to end, not to pin down the numbers. Treat the percentages as provisional: the near-zero C-index entries in particular cannot be told apart from Monte Carlo noise at this replication count. The production run is 100 replications." else ""`

Every column is **% improvement from supplying the real entry time**
(`(phase1 - phase2) / phase1 * 100`). All three metrics are lower-is-better,
so positive = entry time helped; negative = it made that arm/metric worse.

```{r table}
knitr::kable(as.data.frame(summary_tbl, row.names = FALSE))
```

## By metric

```{r plot-setup}
COL_COX  <- "#e6550d"   # same pair used throughout sim_report.R's own plots
COL_RSF  <- "#31a354"
COL_GRID <- "grey85"
COL_REF  <- "grey60"

pct_matrix <- function(field) {
  m <- t(vapply(names(sims), function(nm) {
    res <- res_list[[nm]]
    c(Cox = pct_improve(res, field, "Cox"),
      RSF = pct_improve(res, field, "RSF"))
  }, numeric(2)))
  rownames(m) <- names(sims)
  m
}

plot_metric_bars <- function(field, label) {
  m <- pct_matrix(field)
  ylim <- range(c(0, m), na.rm = TRUE) * 1.15
  barplot(t(m), beside = TRUE, col = c(COL_COX, COL_RSF), border = NA,
          ylim = ylim, las = 1, ylab = "% improvement (phase 1 -> phase 2)",
          main = label, names.arg = rownames(m), cex.names = 0.9)
  abline(h = 0, col = COL_REF, lwd = 1.5)
  grid(nx = NA, ny = NULL, col = COL_GRID, lty = 1)
  box(col = COL_GRID)
  legend("topright", legend = c("Cox", "RSF"), fill = c(COL_COX, COL_RSF),
         border = NA, bty = "n", cex = 0.85)
}
```

Same numbers as the table above, as three small multiples (one axis each —
MAD, C-index, and Brier live on very different scales, so one shared axis
would be misleading). The zero line is the reference: bars crossing below it
are where entry time made that arm/metric *worse*, which only shows up for
RSF's C-index in a few sims.

```{r plot-mad, fig.width=7, fig.height=4}
plot_metric_bars("overall", "MAD vs true S")
```

```{r plot-cindex, fig.width=7, fig.height=4}
plot_metric_bars("cindex", "C-index (ranking)")
```

```{r plot-brier, fig.width=7, fig.height=4}
plot_metric_bars("brier", "Brier / CRPS (calibration)")
```

Raw phase-table numbers (all three metrics, all six sims) are in each sim's
`results_phases.rds` / knitted `stage1_main.html`.
