---
title: "Survival Simulation — Stage 1: Entry Times on Left-Truncated Data"
subtitle: "One sample, two phases: entry zeroed, then entry supplied"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# The experiment

Every replication draws **one** left-truncated sample and runs it through
**two phases**. The rows, the seed, and the hyperparameters are identical
throughout; the only thing that changes between phases is the entry vector
handed to the left-truncation arms.

| | phase 1 — entry zeroed | phase 2 — entry supplied |
|---|---|---|
| **Cox, standard** | `Surv(time, status)` | unchanged |
| **Cox, left-truncation** | `Surv(0, time, status)` | `Surv(entry, time, status)` |
| **RSF, standard** | `rfsrc(...)` | unchanged |
| **RSF, left-truncation** | `rfsrc(..., entry.time = 0)` | `rfsrc(..., entry.time = entry)` |

The standard arms do not read the entry vector at all, so they are fitted once
and carried into both phases unchanged. That makes them the reference line, and
it means each replication needs six fits rather than eight.

## What each phase is for

**Phase 1** should produce four equal numbers. The counting-process Cox with
entry 0 has the same risk sets as the standard fit, and `entry.time = 0` is
documented to reproduce the standard forest bit for bit. Recovering that
equality here confirms the left-truncation code paths are inert when there is no
entry information to use, so anything that happens in phase 2 can be attributed
to the entry times rather than to the modified code being different code.

**Phase 2** supplies the observed entry times to the left-truncation arms. The
standard arms are unchanged by construction, so the reported quantity is the
within-arm shift from phase 1 to phase 2 on identical data.

Both phases sit on left-truncated data. Truncation has already removed subjects
who died before their entry time, and no argument to a model brings them back —
phase 1 is not a truncation-free control, and is not presented as one.

## Why the entry vector changes anything

A subject retained by the filter was under observation only from $R_i$ onward.
A fit that ignores entry places that subject in the risk set over $[0, R_i)$, a
stretch in which it could not have contributed an observed event: the
denominator $Y(t)$ is inflated at early times while the numerator is unchanged.

Replacing $Y_i(t)=\mathbb 1(\tilde T_i \ge t)$ with
$Y_i^{LT}(t)=\mathbb 1(R_i \le t \le \tilde T_i)$ removes those subject-time
contributions. Since $R \perp (\tilde T, X)$ in this DGP, the delayed-entry risk
set at each $t$ is a random subsample of the population risk set at $t$ — the
retained sample is missing people, but what remains is representative of who was
at risk at each time.

## Scoring

All arms are scored against the analytic truth
$$S(t\mid X)=\exp\!\big(-A_0 t^2 e^{\beta^\top X}\big),$$
which truncation leaves unchanged — it is a sampling mechanism, not a change to
the model. Error is computed at every grid point and the headline figure is the
mean of that curve, so the summary number and the curve cannot disagree.

Test covariates are drawn from the **population** distribution rather than the
retained one, since the estimand is the conditional $S(t\mid x)$.

# Setup

This chunk is deliberately **not** cached. Caching it would restore the objects
it creates while silently skipping its `library()` calls, leaving the session
with the engine's functions but none of the packages they need.

```{r libs, cache=FALSE}
source("../_common/sim_engine.R")
source("../_common/sim_report.R")
source("dgp.R")

stopifnot("entry.time" %in% names(formals(rfsrc)))
cat(sprintf("randomForestSRC %s loaded from %s\n",
            packageVersion("randomForestSRC"), path.package("randomForestSRC")))
```

The `stopifnot` guards the most likely silent failure: running against the CRAN
build, where `entry.time` would be absorbed by `...` and both phases would come
out identical for a reason that has nothing to do with the data.

```{r install, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("Shaked06/randomForestSRC@ltrc-rsf")
```

```{r load-setup, cache=FALSE}
setup <- readRDS("setup.rds")
stopifnot(!is.null(setup$ltrc))

M_TEST <- 200L      # test subjects per replication
N_REP  <- 10L      # set to 5 for a fast pass, then 100

str(setup[c("rsf_hp", "ltrc")])
cat(sprintf("\ntime grid: M=%d, range (%.3f, %.3f]\n",
            length(setup$time_grid), min(setup$time_grid), max(setup$time_grid)))
```

Prediction is cheap next to fitting, so `M_TEST` is set well above the number
needed to run: it reduces per-replication noise at almost no cost in wall clock.

# The replication loop

Replications are independent and each seeds itself from `rep_id`, so the
parallel run reproduces a serial one exactly. Workers source the engine and
`dgp.R` rather than receiving exported closures, so each sees what this document
sees. Six fits per replication, of which three are forests.

`N_REP` and `M_TEST` are named in `cache.extra` on purpose. knitr does not track
dependencies across chunks by default, so without this a cached loop would
survive a change to `N_REP` and quietly report the previous run's replication
count.

```{r loop, cache.extra = list(N_REP, M_TEST, tools::md5sum(c("setup.rds", "dgp.R", "../_common/sim_engine.R")))}
res <- run_experiment(
  n_rep   = N_REP,
  setup   = setup,
  sim     = sim,
  sources = c("../_common/sim_engine.R", "dgp.R"),
  m_test  = M_TEST
)

saveRDS(res, "results_phases.rds")
```

# Results

## Phase 1 — are the left-truncation arms inert?

```{r noop, cache=FALSE}
noop_table(res)
```

Both gaps are the largest discrepancy observed across all replications and all
grid points, so a zero here is a statement about every fit in the run.

The reporting chunks are uncached. They cost milliseconds, and caching them
would let a wording change in `sim_report.R` leave stale labels on screen.

## The two phases

```{r phase-table, cache=FALSE}
phase_table(res)
```

Each row is one arm measured in both phases on identical data. The standard rows
carry a shift of exactly zero because those fits never see the entry vector; they
are shown in both columns as the reference line.

## Within-arm shift

Both phases use the same rows and the same seed, so the phase-1 to phase-2
difference is paired within a replication.

```{r paired, cache=FALSE}
paired_shift(res)
```

`n_moved_down` counts replications in which the phase-2 score is the smaller of
the two.

```{r plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res)
```

## C-index (ranking accuracy, training sample)

MAD-vs-truth above checks the predicted *curve* against the true $S(t\mid x)$
on the population test set. This checks the predicted *ranking* of subjects
against who actually failed first, scored on this replication's own real
(time, status, entry) -- RSF's built-in OOB survival curve, Cox's in-sample
fit, both against the same real entry vector regardless of which arm
produced the prediction (see `sim_engine.R` for why). `r res$cindex_label`.

```{r cindex-phase-table, cache=FALSE}
phase_table(res, field = "cindex")
```

```{r cindex-paired, cache=FALSE}
paired_shift(res, field = "cindex")
```

```{r cindex-plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res, field = "cindex")
```

## Brier score / CRPS (calibration, training sample)

C-index above checks *ranking*; this checks *calibration* -- how close the
predicted probability itself is to the observed outcome, at each time point,
integrated into one number (CRPS). Scored the same way as C-index: each
arm's own-data survival curve, against the same real entry vector
regardless of which arm produced the prediction. `r res$brier_label`.

```{r brier-phase-table, cache=FALSE}
phase_table(res, field = "brier")
```

```{r brier-paired, cache=FALSE}
paired_shift(res, field = "brier")
```

```{r brier-plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res, field = "brier")
```

## Time-dependent AUC (ranking at each time point, training sample)

The C-index asks whether the ranking is right *overall*, collapsing every
comparable pair into one number. This asks whether it is right *at each time
point* -- at time `t`, do the subjects who fail by `t` get ranked riskier than
those still event-free, among only those actually under observation then? That
last clause is the one left truncation attacks, which is why a time-resolved
ranking metric earns its place alongside the C-index. Scored on the same data
as both scorers above. `r res$auc_label`.

```{r auc-phase-table, cache=FALSE}
phase_table(res, field = "auc")
```

```{r auc-paired, cache=FALSE}
paired_shift(res, field = "auc")
```

```{r auc-plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res, field = "auc")
```

## Across the time axis

A single summary figure averages over the whole grid and cannot show where along
the time axis the arms separate. The curve below reports the score at each grid
point, averaged over replications.

```{r plot-time, cache=FALSE, fig.width=8, fig.height=5}
invisible(plot_score_by_time(res))
```

Early grid points are worth reading carefully. Few subjects have entered there,
so the delayed-entry risk set is thin, and below the smallest entry time the
hazard is not identified from the data at all. Here $c_L = 0$, so the smallest
entry time among 5000 subjects sits very close to zero and the grid is covered
in practice, but the earliest points remain the least well determined.

# What we have

`results_phases.rds` holds, for `r N_REP` replications: the score for eight
arm-phase combinations, the score-versus-time curve for each, and the phase-1
no-op diagnostics.

The measured quantity is the within-arm shift between the two phases — same
rows, same seed, same hyperparameters, same binary, with the entry vector as the
only difference. The Cox arm applies the same correction through `survival`'s
counting-process code, which is an independent implementation of delayed entry.

Scoring against the analytic truth rather than out-of-bag error avoids the
delayed-entry C-index and Brier score, which are not implemented in the fork.

## Reusing this format

The machinery lives in `../_common/sim_engine.R`: estimators, scoring, the two
phases, the parallel loop, and the tables and plots above. A new simulation
supplies a `dgp.R` defining `sim` with `draw_train`, `draw_test`, and `truth`,
then reuses this document's structure unchanged. Datasets with no analytic truth
can pass a different scorer to `run_experiment`; a scorer needs a `label` and a
`by_time(Shat, ctx)` function.
