---
title: "Survival Simulation 04 — Stage 0: Setup (DGP, Calibration, Tuning)"
subtitle: "RIST paper Scenario 2 (nonlinear, PH-violating link), with left truncation"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# Background

This is the fourth simulation in this project, reusing the two-stage format
established by simulations 01-03 — same shared engine
(`../_common/sim_engine.R`), same reporting (`../_common/sim_report.R`) — on
a fourth DGP.

The DGP is Scenario 2 from Zhu and Kosorok (2012), "Recursively Imputed
Survival Trees," Section 4.1
(`../../papers/Recursively Imputed Survival Trees.pdf`, p. 334): a
nonlinear link deliberately built to violate the proportional-hazards
assumption, unlike simulation 03's Scenario 1 (exactly Cox) and closer in
spirit to simulation 02 (also a deliberate PH violation, but on a different
DGP).

## What is taken from the paper, and what we add

Taken from RIST's Scenario 2 (unchanged):

- $p=10$ iid covariates. The paper does not state a range for "uniform
  distributed covariates"; we read this as $\mathrm{Unif}(0,1)$, the
  standard default, consistent with the link function's $|x_2-0.5|$ term
  being centered where you'd expect. Only $x_1,x_2,x_3$ enter the model;
  $x_4,\dots,x_{10}$ are pure noise.
- $\mu(X) = \sin(x_1\pi) + 2|x_2-0.5| + x_3^3$
- $T\sim\mathrm{Exponential}(\text{mean}=\mu(X))$
- $C\sim\mathrm{Uniform}(0,U)$, a single shared upper bound (not
  per-subject) — a genuinely different censoring *shape* from Scenario 1's
  Exponential, and the reason the left-truncation layer below also stays
  Uniform rather than reusing simulations 01-03's Exponential recipe.
- target censoring rate $\approx24\%$

No separate administrative cutoff is used here: $C\le U$ always, so the
censoring distribution's own upper bound already caps every observed time —
unlike simulations 01-03, there is no extra `pmin(., tau)`.

Added here, present in neither this scenario nor the source paper (Scenario
2 has no left truncation):

- entry time $R\sim\mathrm{Unif}(c_L,c_U)$, retention filtered on
  $R\le\tilde T$ — the same mechanism simulations 01-03 use.
- residual censoring after entry, but keeping the paper's own **Uniform**
  shape: $C = R + \mathrm{Unif}(0, w_{LT})$, rather than switching to
  Exponential. Using Exponential here would silently erase the one thing
  that distinguishes Scenario 2's censoring design from Scenario 1's.
- $c_U$, $U$ (reference), and $w_{LT}$ are all calibrated by root-find
  against a target rate, following this project's established pattern,
  rather than taken directly from the paper's fixed $\tau=6$. RIST's
  $\approx24\%$ is treated as the design target for both $U$ and $w_{LT}$ —
  the same resolution used for simulation 03's censoring discrepancy. In
  practice both calibrated values land close to the paper's own
  $\tau=6$ (a check worth re-running here, see below), which is a useful
  sanity check on the calibration itself.
- sample sizes, matching simulations 01-03 rather than the paper's own
  $n=200$: a population pool of $N_{\text{pool}} = 50{,}000$ filtered down
  to a retained training set of $N_{\text{ltrc}} = 5{,}000$ per replication.
- RSF hyperparameters: when not OOB-tuned, fall back to RIST's own Section
  4.2 recipe ($K=\lfloor\sqrt p\rfloor$, $n_{\min}=6$), same as simulation
  03 but recomputed for $p=10$: $\lfloor\sqrt{10}\rfloor=3$.

```{r libs}
source("dgp.R")
suppressPackageStartupMessages({
  library(survival)
  library(randomForestSRC)
})
```

Outputs of this stage, all written to `setup.rds`:

1. **$U$** — right-censoring upper bound for the untruncated reference DGP,
   calibrated to $\approx24\%$ censoring.
2. **$c_U$** — entry-window upper bound, calibrated so that $\approx\tfrac13$
   of the population is truncated away.
3. **$w_{LT}$** — censoring width under truncation, calibrated to
   $\approx24\%$ censoring among the *retained* subjects.
4. **Time grid** $t_1,\dots,t_M$.
5. **RSF hyperparameters**.

## Ordering

As in simulations 01-03, the time grid and RSF tuning are derived from a
**left-truncated** pilot sample, so the sequence runs
$U \to c_U \to w_{LT} \to$ pilot sample $\to$ grid $\to$ tuning.

```{r params}
MASTER_SEED   <- 20250214L   # distinct from simulations 01-03's seeds

# --- calibration targets ---
N_PILOT       <- 50000L
TARGET_CENSOR <- 0.24        # RIST Scenario 2's stated ~24%, as a design target
TARGET_TRUNC  <- 1/3         # same target as simulations 01-03
M_TIMES       <- 100L
TIME_QUANTILE <- 0.90

# --- left-truncation settings ---
N_POOL        <- 50000L
N_LTRC        <- 5000L       # matches simulations 01-03's run size
C_L           <- 0

# --- RSF tuning ---
# See simulation 01's Stage 0 for why RUN_TUNE defaults off. RUN_TUNE=FALSE
# falls back to RIST's own RSF settings (Section 4.2): K = floor(sqrt(p)),
# nmin = 6 -- recomputed here for p=10 rather than reusing simulation 03's
# p=25 value.
RUN_TUNE      <- FALSE
RSF_MTRY_DEFAULT     <- as.integer(floor(sqrt(P)))   # RIST Sec. 4.2: K = floor(sqrt(p))
RSF_NODESIZE_DEFAULT <- 6L                            # RIST Sec. 4.2: n_min = 6
```

# Censoring upper bound $U$ (untruncated reference)

Because $C \perp \tilde T$, the censoring rate integrates out in closed form
given the pilot's $\tilde T$ values,
$$g(U)=\frac1n\sum_i \frac{\min(\tilde T_i, U)}{U},$$
which is monotone decreasing in $U$ (a wider censoring window pushes $C$
higher on average, so it is less likely to land below $\tilde T$).
`uniroot` therefore returns a unique root with no sampling noise.

This parameter belongs to the untruncated DGP. Stage 1 fits only truncated
data, so it is carried in `setup.rds` for reference rather than consumed
downstream.

```{r censoring-U}
set.seed(MASTER_SEED)

X_pilot      <- sample_X(N_PILOT)
Ttilde_pilot <- sample_event_times(X_pilot)

U_hat <- uniroot(
  f        = function(u) censoring_rate(u, Ttilde_pilot) - TARGET_CENSOR,
  interval = c(1e-3, 50 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

chk_df     <- make_survival_sample(N_PILOT, U_hat)
emp_censor <- mean(chk_df$status == 0)

cat(sprintf("U = %.4f  (paper's own tau = 6, for comparison)\n", U_hat))
cat(sprintf("  censoring  analytic = %.3f | empirical = %.3f\n",
            censoring_rate(U_hat, Ttilde_pilot), emp_censor))
```

# Entry window $c_U$

Integrating out $R \sim \mathrm{Unif}(c_L, c_U)$ given the pilot's $\tilde T$:
$$\mathbb P(R > \tilde T_i) =
\begin{cases}
1 & \tilde T_i < c_L\\[2pt]
\dfrac{c_U - \tilde T_i}{c_U - c_L} & c_L \le \tilde T_i \le c_U\\[6pt]
0 & \tilde T_i > c_U
\end{cases}$$
increasing in $c_U$.

```{r cU}
cU_hat <- uniroot(
  f        = function(cu) trunc_rate(cu, Ttilde_pilot, C_L) - TARGET_TRUNC,
  interval = c(C_L + 1e-6, 50 * max(Ttilde_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("cU = %.4f  (entry window: Unif(%.2f, %.2f))\n", cU_hat, C_L, cU_hat))
cat(sprintf("  truncated away  analytic = %.3f  (target %.3f)\n",
            trunc_rate(cU_hat, Ttilde_pilot, C_L), TARGET_TRUNC))
```

# Censoring under truncation, $w_{LT}$

Calibrated on the **retained** subjects, since that is the population the
models see. With $C_i = R_i + W_i$ and $W_i\sim\mathrm{Unif}(0,w_{LT})$, a
subject is censored iff $W_i < \tilde T_i - R_i$, so
$$\mathbb P(C_i < \tilde T_i) = \frac{\min(\tilde T_i - R_i,\, w_{LT})}{w_{LT}},$$
decreasing in $w_{LT}$ for the same reason as $U$ above.

```{r w-lt}
set.seed(MASTER_SEED + 2L)

R_pilot  <- runif(N_PILOT, C_L, cU_hat)
kept     <- R_pilot <= Ttilde_pilot
Tk_pilot <- Ttilde_pilot[kept]
Rk_pilot <- R_pilot[kept]

w_lt_hat <- uniroot(
  f        = function(w) censoring_rate_lt(w, Tk_pilot, Rk_pilot) - TARGET_CENSOR,
  interval = c(1e-6, 50 * max(Tk_pilot - Rk_pilot)),
  tol      = 1e-6
)$root

cat(sprintf("retained %d / %d pool subjects (%.1f%%)\n",
            sum(kept), N_PILOT, 100 * mean(kept)))
cat(sprintf("w_LT = %.4f  (paper's own tau = 6, for comparison)\n", w_lt_hat))
cat(sprintf("  censoring (retained)  analytic = %.3f\n",
            censoring_rate_lt(w_lt_hat, Tk_pilot, Rk_pilot)))
```

Both $U$ and $w_{LT}$ landing close to the paper's own $\tau=6$ is a useful
check on the calibration: hitting the paper's stated $\approx24\%$ target
recovers something close to the scale the paper itself used, rather than
drifting off to an implausible value.

# The left-truncated pilot sample

One sample at the Stage 1 run size, used for both the grid and the tuning.

```{r ltrc-pilot}
set.seed(MASTER_SEED + 3L)
df_pilot <- make_ltrc_sample(N_LTRC, N_POOL, w_lt_hat, cU_hat, C_L)

cat(sprintf("LTRC pilot: n=%d,  censoring=%.3f,  entry <= time on all rows: %s\n",
            nrow(df_pilot), mean(df_pilot$status == 0),
            all(df_pilot$entry <= df_pilot$time)))
```

# Time grid

As in simulations 01-03, taken from the **left-truncated** pilot — the
regime Stage 1 fits — up to the 90th percentile of observed follow-up.

```{r grid}
upper     <- as.numeric(quantile(df_pilot$time, TIME_QUANTILE))
time_grid <- seq(0, upper, length.out = M_TIMES + 1L)[-1L]

cat(sprintf("time grid: M=%d, range (%.3f, %.3f]\n",
            length(time_grid), min(time_grid), max(time_grid)))
cat(sprintf("cU = %.3f sits at grid position %d of %d\n",
            cU_hat, sum(time_grid < cU_hat) + 1L, length(time_grid)))
```

# RSF hyperparameters

Same choices as simulations 01-03 when tuning does run: on the
left-truncated pilot, and entry-agnostic. `RUN_TUNE` gates the search; with
`RUN_TUNE <- FALSE`, `rsf_hp` falls back to RIST's Section 4.2 recipe for
$p=10$: `mtry = floor(sqrt(10)) = 3`, `nodesize = 6`.

```{r rsf-tune, eval=RUN_TUNE}
rsf_tune <- tune(
  Surv(time, status) ~ .,
  data      = df_pilot[, c(grep("^x", names(df_pilot), value = TRUE),
                           "time", "status")],
  ntree.try = 300,
  do.best   = FALSE,
  seed      = MASTER_SEED
)
```

```{r rsf-hp}
rsf_hp <- if (RUN_TUNE) {
  list(ntree    = 1000L,
       mtry     = as.integer(rsf_tune$optimal["mtry"]),
       nodesize = as.integer(rsf_tune$optimal["nodesize"]))
} else {
  list(ntree = 500L, mtry = RSF_MTRY_DEFAULT, nodesize = RSF_NODESIZE_DEFAULT)
}

cat(sprintf("RSF hyperparameters (%s): mtry=%d, nodesize=%d, ntree=%d\n",
            if (RUN_TUNE) "OOB-tuned" else "fixed, RIST Sec. 4.2 recipe",
            rsf_hp$mtry, rsf_hp$nodesize, rsf_hp$ntree))
```

# Sanity checks

**(a) Inverse-transform formula.** If sampling is correct then
$\tilde T_i / \mu(X_i) = -\ln U_i \sim \mathrm{Exp}(1)$, so mean and
variance should both be $\approx 1$.

```{r check-inv}
exp1_chk <- Ttilde_pilot / mu_link(X_pilot)
cat(sprintf("Exp(1) check: mean=%.3f (~1), var=%.3f (~1)\n",
            mean(exp1_chk), var(exp1_chk)))
```

**(b) Monotonicity of $S$.**

```{r check-surv}
s_chk <- as.vector(true_survival_matrix(X_pilot[1, , drop = FALSE], time_grid))
cat(sprintf("S range [%.3f, %.3f]; monotone decreasing: %s\n",
            min(s_chk), max(s_chk), all(diff(s_chk) <= 0)))
```

**(c) $\mu(X)$ range.** Should stay in $[0,3]$ and away from 0, given
$x_1,x_2,x_3\sim\mathrm{Unif}(0,1)$.

```{r check-mu}
mu_chk <- mu_link(X_pilot)
cat(sprintf("mu range [%.4f, %.4f], any non-positive: %s\n",
            min(mu_chk), max(mu_chk), any(mu_chk <= 0)))
```

**(d) Selection is present.** Retained subjects survived to their own entry
time, so — as in simulation 03, and unlike simulation 01 — subjects with
**larger** $\mu(X)$ (longer $\tilde T$ on average) should be over-represented
among the retained.

```{r check-selection}
mu_pool <- mu_link(X_pilot)
cat(sprintf("mean mu(X)   pool = %.4f | retained = %.4f\n",
            mean(mu_pool), mean(mu_pool[kept])))
```

Note that this shift does not by itself bias the target. The estimand is
the conditional $S(t\mid x)$, and Stage 1 evaluates on covariates drawn
from the population distribution. What matters is the risk set: under
$R \perp (\tilde T, X)$ the delayed-entry risk set at each $t$ is a random
subsample of the population risk set at $t$.

**(e) Pool headroom.** The pool must retain at least `N_LTRC` subjects.

```{r check-pool}
cat(sprintf("need %d, expect ~%.0f retained from a pool of %d\n",
            N_LTRC, (1 - TARGET_TRUNC) * N_POOL, N_POOL))
```

# Saving

```{r save}
setup <- list(
  rsf_hp    = rsf_hp,
  time_grid = time_grid,
  cens_U    = U_hat,
  ltrc      = list(
    cL     = C_L,
    cU     = cU_hat,
    w_lt   = w_lt_hat,
    n_pool = N_POOL,
    n_ltrc = N_LTRC
  ),
  meta      = list(
    n_pilot       = N_PILOT,
    target_censor = TARGET_CENSOR,
    emp_censor    = emp_censor,
    target_trunc  = TARGET_TRUNC,
    emp_retained  = mean(kept),
    emp_censor_lt = mean(df_pilot$status == 0),
    grid_source   = "ltrc",
    tuned_on      = if (RUN_TUNE) "ltrc" else "fixed (RIST Sec. 4.2 recipe, RUN_TUNE=FALSE)",
    master_seed   = MASTER_SEED,
    source        = "Recursively Imputed Survival Trees.pdf, Scenario 2 (p.334)"
  )
)
saveRDS(setup, "setup.rds")
cat("setup.rds saved — ready for Stage 1.\n")
```
