---
title: "Survival Simulation 04 — Stage 1: Entry Times on RIST Scenario 2"
subtitle: "One sample, two phases: entry zeroed, then entry supplied"
author: "Shaked"
date: "`r format(Sys.Date(), '%d/%m/%Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  cache   = TRUE,
  fig.align = "center"
)
options(digits = 4)
```

# The experiment

Same design as simulations 01-03's Stage 1, run against a fourth DGP. Every
replication draws **one** left-truncated sample and runs it through **two
phases**. The rows, the seed, and the hyperparameters are identical
throughout; the only thing that changes between phases is the entry vector
handed to the entry-aware arms.

| | phase 1 — entry zeroed | phase 2 — entry supplied |
|---|---|---|
| **Cox, standard** | `Surv(time, status)` | unchanged |
| **Cox, left-truncation** | `Surv(0, time, status)` | `Surv(entry, time, status)` |
| **RSF, standard** | `rfsrc(...)` | unchanged |
| **RSF, left-truncation** | `rfsrc(..., entry.time = 0)` | `rfsrc(..., entry.time = entry)` |

The standard arms do not read the entry vector at all, so they are fitted
once and carried into both phases unchanged.

## What is different from simulations 01-03

The DGP is Scenario 2 from the RIST paper's simulation study
(`../../papers/Recursively Imputed Survival Trees.pdf`, Section 4.1,
p. 334), not reproduced here — see `dgp.R` and this simulation's Stage 0
for the model and the closed-form truth. Unlike simulation 03 (Scenario 1,
exactly Cox), this DGP is built to **violate proportional hazards**: with
$$\mu(X) = \sin(x_1\pi) + 2|x_2-0.5| + x_3^3, \qquad T\sim\mathrm{Exponential}(\text{mean}=\mu(X)),$$
the hazard is $1/\mu(X)$ — constant over time for a given subject, but not
a simple multiplicative effect of the covariates, so a Cox fit is
misspecified by construction. In that sense it plays the same role
simulation 02 plays relative to simulation 01, but on RIST's own DGP rather
than the flexible-hazard paper's. Censoring is also a different *shape*
here — Uniform rather than Exponential — carried through into the added
left-truncation layer as well (see Stage 0).

Left truncation is not part of the RIST paper. It is added here as in
simulations 01-03: entry times $R\sim\mathrm{Unif}(c_L,c_U)$ independent of
$(\tilde T,X)$, retention filtered on the true event time, and residual
censoring after entry — Uniform-shaped here, so `entry <= time` holds on
every row.

## What each phase is for

**Phase 1** should produce four equal numbers, for the same reason as
simulations 01-03: the counting-process Cox with entry 0 has the same risk
sets as the standard fit, and `entry.time = 0` is documented to reproduce
the standard forest bit for bit.

**Phase 2** supplies the observed entry times. The reported quantity is the
within-arm shift from phase 1 to phase 2 on identical data.

## Scoring

All arms are scored against the closed-form truth
$$S(t\mid X)=\exp\!\big(-t / \mu(X)\big),$$
which truncation leaves unchanged. Test covariates are drawn from the
**population** distribution, since the estimand is the conditional
$S(t\mid x)$.

# Setup

This chunk is deliberately **not** cached — caching it would restore the
objects it creates while silently skipping its `library()` calls.

```{r libs, cache=FALSE}
source("../_common/sim_engine.R")
source("../_common/sim_report.R")
source("dgp.R")

stopifnot("entry.time" %in% names(formals(rfsrc)))
cat(sprintf("randomForestSRC %s loaded from %s\n",
            packageVersion("randomForestSRC"), path.package("randomForestSRC")))
```

```{r install, eval=FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
remotes::install_github("Shaked06/randomForestSRC@ltrc-rsf")
```

```{r load-setup, cache=FALSE}
setup <- readRDS("setup.rds")
stopifnot(!is.null(setup$ltrc))

M_TEST <- 200L      # test subjects per replication
N_REP  <- 10L       # set lower for a fast pass

str(setup[c("rsf_hp", "ltrc")])
cat(sprintf("\ntime grid: M=%d, range (%.3f, %.3f]\n",
            length(setup$time_grid), min(setup$time_grid), max(setup$time_grid)))
```

# The replication loop

Same mechanics as simulations 01-03: replications are independent, each
seeds itself from `rep_id`, and workers source the engine and `dgp.R`
rather than receiving exported closures.

`N_REP` and `M_TEST` are named in `cache.extra` so a cached loop cannot
silently survive a change to either.

```{r loop, cache.extra = list(N_REP, M_TEST, tools::md5sum(c("setup.rds", "dgp.R", "../_common/sim_engine.R")))}
res <- run_experiment(
  n_rep   = N_REP,
  setup   = setup,
  sim     = sim,
  sources = c("../_common/sim_engine.R", "dgp.R"),
  m_test  = M_TEST
)

saveRDS(res, "results_phases.rds")
```

# Results

## Phase 1 — are the left-truncation arms inert?

```{r noop, cache=FALSE}
noop_table(res)
```

The reporting chunks below are uncached: they cost milliseconds, and caching
them would let a wording change in `sim_report.R` leave stale labels on
screen.

## The two phases

```{r phase-table, cache=FALSE}
phase_table(res)
```

## Within-arm shift

```{r paired, cache=FALSE}
paired_shift(res)
```

`n_moved_down` counts replications in which the phase-2 score is the smaller
of the two.

```{r plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res)
```

## C-index (ranking accuracy, training sample)

MAD-vs-truth above checks the predicted *curve* against the true $S(t\mid x)$
on the population test set. This checks the predicted *ranking* of subjects
against who actually failed first, scored on this replication's own real
(time, status, entry) -- RSF's built-in OOB survival curve, Cox's in-sample
fit, both against the same real entry vector regardless of which arm
produced the prediction (see `sim_engine.R` for why). `r res$cindex_label`.

```{r cindex-phase-table, cache=FALSE}
phase_table(res, field = "cindex")
```

```{r cindex-paired, cache=FALSE}
paired_shift(res, field = "cindex")
```

```{r cindex-plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res, field = "cindex")
```

## Brier score / CRPS (calibration, training sample)

C-index above checks *ranking*; this checks *calibration* -- how close the
predicted probability itself is to the observed outcome, at each time point,
integrated into one number (CRPS). Scored the same way as C-index: each
arm's own-data survival curve, against the same real entry vector
regardless of which arm produced the prediction. `r res$brier_label`.

```{r brier-phase-table, cache=FALSE}
phase_table(res, field = "brier")
```

```{r brier-paired, cache=FALSE}
paired_shift(res, field = "brier")
```

```{r brier-plot-box, cache=FALSE, fig.width=7.5, fig.height=4.5}
plot_phases_box(res, field = "brier")
```

## Across the time axis

```{r plot-time, cache=FALSE, fig.width=8, fig.height=5}
invisible(plot_score_by_time(res))
```

# What we have

`results_phases.rds` holds, for `r N_REP` replications, the score for eight
arm-phase combinations, the score-versus-time curve for each, and the
phase-1 no-op diagnostics — same structure as simulations 01-03, on a DGP
where the Cox arms are misspecified by construction. Comparing this to
simulation 03 (same paper, exactly-Cox Scenario 1) isolates the effect of
proportional-hazards violation on top of the same left-truncation question;
comparing it to simulation 02 (a different paper's PH-violating DGP) checks
whether that effect is specific to one DGP's way of violating PH.
